---
title: "Low vs high differential abundance"
author: "Wanxin_Lai"
date: "2024-03-25"
output:
  rmarkdown::html_document:
    theme: lumen
    highlight: zenburn
    df_print: default
    toc: yes
    toc_float: TRUE
    keep_md: no
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
setwd("C:/Users/awala/OneDrive - Norwegian University of Life Sciences/Documents/Small_Cow/MAG/")
```

```{r, message = FALSE, warning=FALSE}
library(tidyverse)
library(ggtree)
library(phytools)
library(scales)
library(ggnewscale)
library(ggh4x)
library(gridExtra)
library(ggtreeExtra)
library(DT)
library(pheatmap)
```

```{r message = FALSE}
#Percentage of genomes covered by reads
CF <- read_tsv("./Abundance.coverM_Fraction.tsv")

# Remove T6 samples
# Modify abundance data
covered_fraction <- CF %>%
  select(Genome,matches("Covered Fraction")) %>%
  filter(Genome != "unmapped")

new_column_names <- covered_fraction %>%
  select(-1) %>%  # Exclude the first column
  names() %>%
  stringr::str_replace(".R1\\.fq\\.gz Covered Fraction", "")

coverage_table <- covered_fraction %>%
  rename_with(~ new_column_names,-1) %>% 
  column_to_rownames("Genome") %>% 
  select(-matches("T6"))
```

Genomes that have less than 30% of their length covered by reads are turned into zeros to account for the random allocation of reads across genomes due to mapping heuristics.

```{r message = FALSE}
count <- read_tsv("./misc/Counts.tsv", col_names=T)

#Remove T6 samples
count_table <- count %>% select(-matches("T6"))

#rearrange row and col order based on count table 
row_order <- count_table$Genome
col_order <- colnames(count_table[,c(-1,-2,-3)])

coverage_table.sort <- coverage_table[row_order,col_order]

#filter low-coverage reads
min_coverage=0.3

# if not zero counts, assign 1
count_table_cov <- coverage_table.sort %>% 
  mutate(across(everything(), ~ ifelse(. > min_coverage, 1, 0))) %>% 
  map2_df(., count_table[, c(-1,-2,-3)], ~ .x * .y) %>% # df x times df y, get the actual count if cov is 1 (>0.3)
  as.data.frame() %>%
  select(-c("D55T5", "D67T5"))

rownames(count_table_cov) <- rownames(coverage_table.sort)
```

```{r Phylum colour scheme, message=FALSE}

phylum_colour <- data.frame(Phylum = sort(unique(count$Phylum)), color = hue_pal()(12))

mags_phylum_colour <- count %>% 
  select("Genome", "classification", "Phylum") %>% 
  left_join(phylum_colour, by = join_by(Phylum == Phylum))
```


#### Normalise read counts into genome counts
The relative representation of each genome depends on its size. 
```{r message = FALSE}
#load MAGs table
mags_table <- read_tsv("./results_ac2/checkm2_quality_report.tsv", col_names = T) %>% 
  column_to_rownames(var = "Name")

# Transform reads into genome counts 
# modifying the reads-counts based on the normalisation factor, which is the average genome size (mag_size)
genome_read_sizes <- mags_table[rownames(count_table_cov),] %>%
    select(Genome_Size) %>%
    mutate(mag_size = Genome_Size / 150) %>%
    pull()

# Compare the size of each genome to the average genome size in the dataset
# 1 is row, dvd the read size to the row of count_table_cov.
count_table_cov_size <- sweep(count_table_cov, 1, genome_read_sizes, "/")

# Do the same normalisation for the unfiltered reads counts

raw_count_table_cov_size <- count_table[,-c(2,3)] %>% 
  column_to_rownames(var = "Genome") %>% 
  sweep(., 1, genome_read_sizes, "/")

```
 
##### Visualise the relative MAG abundances per sample after filtering the low coverage genome-counts. 

```{r import mashtree file, message = FALSE}

tree <- read.tree("./results_ac2/mashtree/mashtree.newick")

```

##### Generates a count table in which the scale is log10-transformed.
```{r vertical tree, message = FALSE}
phylum_colour <- count %>% 
  select(Genome, Phylum) %>% 
  column_to_rownames("Genome")

# Generate phylum color heatmap
heatmap <- mags_phylum_colour %>% arrange(match(Genome, tree$tip.label)) %>%
  select(Genome,Phylum) %>%
    mutate(Phylum = factor(Phylum, levels = unique(Phylum))) %>%
    column_to_rownames(var = "Genome")

# create ultrametric tree
vertical_tree <- force.ultrametric(tree,method="nnls") %>%
        ggtree(., size = 0.1)

#first test
vertical_tree <- gheatmap(vertical_tree, heatmap, width = 0.05, offset = 0, colnames = FALSE)

#Reset fill scale
vertical_tree <- vertical_tree + new_scale_fill()

count_table_cov_size <- count_table_cov_size %>% 
  rownames_to_column(var = "Genome") %>%  
  arrange(match(Genome, tree$tip.label)) %>% 
  column_to_rownames("Genome")
```

```{r, message = FALSE, warning=FALSE, fig.width= 9, fig.height=7}

#Add counts
vertical_tree_count_tbl <- gheatmap(vertical_tree, 
                          log10(count_table_cov_size), 
                          offset = 0.2,
                          colnames_angle=90, 
                          font.size=3, 
                          colnames=TRUE,
                          colnames_position="top", 
                          colnames_offset_y = 15
                          ) +
  vexpand(.08) +
  coord_cartesian(clip = "off") +
  scale_fill_gradient(low = "white", high = "steelblue", na.value="white", name = "log10(RA)")

vertical_tree_count_tbl
```

After applying coverage filtering, the retained dataset appears sparse, with numerous phyla showing predominantly NA (white) counts.

##### Classic phylogenomics tree of 700 MAGs (raw MAGs count table) 
```{r, fig.height=30, fig.width= 10}
# Make a back up with tip labels being replaced
tree.bk <- tree

# Arrange colors alphabetically
colors_alphabetic <- mags_phylum_colour %>% arrange(match(Genome, tree.bk$tip.label)) %>%
  select(Phylum, color) %>%
    unique() %>%
    arrange(Phylum) %>%
    select(color) %>%
    pull()

#Sort phylum table by tree topology and add colors
mags_phylum_genomes_sorted <- mags_phylum_colour %>%
  arrange(match(Genome, tree.bk$tip.label))

tree.bk$tip.label <- unlist(lapply(mags_phylum_genomes_sorted[,2], as.character))


#Generate Phylum tree image

plot(tree.bk, tip.color = mags_phylum_genomes_sorted$color, cex = 0.4)

```

```{r, eval=FALSE}
png("phylum_tree.png",width=800,height=2000)
ggtree(tree.bk) + geom_tiplab(size = 1.5, color = mags_phylum_genomes_sorted$color) + xlim(NA,2) + theme(legend.position='right') + geom_treescale(fontsize=3.5, linesize=0.5, offset=0.5)
dev.off()

pdf("phylum_tree.pdf",width=27,height=35)
ggtree(tree.bk) + geom_tiplab(size = 1.5, color = mags_phylum_genomes_sorted$color, align = F) + xlim(NA,2) + theme(legend.position='right') + geom_treescale(fontsize=3.5, linesize=0.5, offset=0.5)
dev.off()
```

##### Create Stacked barplot (filtered coverage)
```{r Load metadata}
T16meta <- read.table("../counttable/CP958c_T1-T6-metadata.txt", header = T, sep = "\t", row.names = 1)
#metadata <- data.frame(row.names = T16meta[,1], T16meta[2:ncol(T16meta)])

# include the recent split from Velma and Carl metadata
vsplit <- read.table("../counttable/metadata_v1.5 1.tsv", header = T, sep = "\t", text = "animal", colClasses = "character")
vsplit$animal <- paste("A", vsplit$animal, sep="")

# concatenate two metadata into 53 samples
metadata <- T16meta %>%
  rownames_to_column("SampleID") %>% 
  left_join(vsplit, by = c("Animal" = "animal"))

metadata <- data.frame(row.names = metadata[,2], metadata[,c(3,4,5,6,7,29)])

metadata <- metadata %>% filter(!str_detect(Time, "T6"))
```

#### Table for taxonomy barplot 
```{r}
count_table_cov_size_pivot <- count_table_cov_size %>%
  rownames_to_column("Genome") %>%
  
  #apply TSS 
  mutate_at(vars(-Genome),~./sum(.)) %>% 
  
  #reduce to minimum number of columns
  pivot_longer(-Genome, names_to = "sample", values_to = "count") %>%
  
  left_join(., mags_phylum_genomes_sorted, by = join_by(Genome == Genome)) %>%
  
   #sort phyla by taxonomy
  mutate(Phylum = fct_relevel(Phylum)) %>%
  
  left_join(., (metadata %>% rownames_to_column(var = "sample") %>% 
            select(sample, Time, Emission, Breed, vsplit_old)), by = join_by("sample" == "sample")) %>% 
   filter(!is.na(count))

datatable(count_table_cov_size_pivot)
```

#### Stacked barplot per individual

```{r Stacked barplot per individual, fig.width= 12}

# Facet groups for coloring
facet_boxes <- metadata %>%
  #filter(!is.na(count)) %>%
  #filter(!is.na(Animal)) %>%
  #filter(Time %in% c("T1","T2","T3","T4","T5")) %>%
    distinct(Emission, Breed, Time) %>%
    arrange(Emission, Time)

### This part is a terrible programming for the stacked barplot because ===================
### case_match() doesnt work on repeats ===================================================

# create a nested object (stripped bars)
strip_background <- strip_nested(background_y = elem_list_rect(fill = c(
        # level1 colors
        case_match(
          c(unique(facet_boxes$Emission), unique(facet_boxes$Breed), unique(facet_boxes$Breed), unique(facet_boxes$Time), unique(facet_boxes$Time), unique(facet_boxes$Time), unique(facet_boxes$Time)[-2]),
          "high" ~ "brown",
          "low" ~ "orange",
          "Aberdeen Angus X" ~ "turquoise",
          "Luing" ~ "yellow",
          "Aberdeen Angus X" ~ "turquoise",
          "Luing" ~ "yellow",
          "T1" ~ "#A3A500",
          "T2" ~ "#a7ab7e",
          "T3" ~ "#adcde0",
          "T4" ~ "#00BF7A",
          "T5" ~ "#00C1A9",
          "T1" ~ "#A3A500",
          "T2" ~ "#a7ab7e",
          "T3" ~ "#adcde0",
          "T4" ~ "#00BF7A",
          "T5" ~ "#00C1A9",
          "T1" ~ "#A3A500",
          "T2" ~ "#a7ab7e",
          "T3" ~ "#adcde0",
          "T4" ~ "#00BF7A",
          "T5" ~ "#00C1A9",
          "T1" ~ "#A3A500",
          #"T2" ~ "#a7ab7e",
          "T3" ~ "#adcde0",
          "T4" ~ "#00BF7A",
          "T5" ~ "#00C1A9",
          .default = "grey"
        )
    )))

#=========================================================================================

ggplot((count_table_cov_size_pivot %>%  
          filter(count > 0)) , aes(x=count,y=sample, fill=Phylum, group=Phylum)) + #grouping enables keeping the same sorting of taxonomic units
  
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    labs(x = "Relative abundance", y ="Samples") +
    facet_nested(Emission + Breed + Time ~ .,  scales="free", space = "free_y", strip = strip_background) + #facet per day and treatment
    guides(fill = guide_legend(ncol = 1)) +
    scale_x_continuous(expand = c(0.001, 0.001)) +
    theme(axis.text.y = element_text(),
          axis.title.x = element_blank(),
          strip.text.y = element_text(angle = 0),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
          panel.spacing = unit(0, "lines"),
          legend.position = "right")
```



### MAGs quality visualisation 
#### (Filtered coverage count data)
```{r}
# Get Taxa info from metaP

SamplesNames <- readRDS("./01_map_proteome2samples_c6.rds (1).gz") %>%
  select(sample,matches("tax")) %>%
  distinct(sample,.keep_all = T)


SampleTax <- SamplesNames %>% 
  column_to_rownames("sample") %>% 
  setNames(as_tibble(colnames(.)[2:9]) %>%
             mutate(NEW=c("Domain","Kingdom","Phylum","Class","Order","Family","Genus","Species")) %>%
             select(NEW) %>%
             deframe()) %>%
  rownames_to_column("sample") %>%
  mutate_at(.vars=vars(Species),
            .funs = ~
              str_remove(.,
                         pattern="NA")) %>%
  mutate_at(.vars=vars(Species),
            .funs = ~
              str_remove(.,
                         pattern = ".*[[:space:]]")) %>%
  mutate_at(.vars=vars(Species),
            .funs = ~
              str_remove(.,
                         pattern = "(?<=sp)\\d.*")) %>%
  mutate_at(.vars = vars(Domain),
            .funs = ~
              str_remove_all(.,
                             pattern = ".*d__")) %>%
  mutate_at(.vars = vars(Phylum),
            .funs = ~
              str_remove_all(.,
                             pattern = ".*p__")) %>%
  mutate_at(.vars = vars(Phylum),
            .funs = ~
              str_remove_all(.,
                             pattern = "_..*")) %>% 
  mutate_at(.vars = vars(Class),
            .funs = ~
              str_remove_all(.,
                             pattern = ".*c__")) %>%
  mutate_at(.vars = vars(Order),
            .funs = ~
              str_remove_all(.,
                             pattern = ".*o__")) %>%
  mutate_at(.vars = vars(Family),
            .funs = ~
              str_remove_all(.,
                             pattern = ".*f__")) %>%
  mutate_at(.vars = vars(Genus),
            .funs = ~
              str_remove_all(.,
                             pattern = ".*g__")) %>%
  mutate_at(.vars = vars(Genus),
            .funs = ~
              str_remove_all(.,
                             pattern = "_..*")) %>%
  mutate_at(.vars = vars(Species),
            .funs = ~
              str_remove_all(.,
                             pattern = ".*s__")) %>% 
  mutate(Family = if_else(Family == "NA", Order, Family)) %>% 
  mutate(Genus = if_else(Genus == "NA", Family, Genus))

#convert empty string to sp
SampleTax$Species <- replace(SampleTax$Species, SampleTax$Species == "", "sp")


```


#### MAG quality table
```{r message = FALSE, warning=FALSE}
mags_details <- mags_table %>%
  rownames_to_column(var = "Genome") %>% 
  left_join(., SampleTax, by = join_by(Genome == sample)) %>% 
  select(c(Genome,Domain,Phylum,Genus,Species,Completeness,Contamination,Genome_Size)) %>%
  
  #change mags_size to MBs
  mutate(mag_size=round(Genome_Size/1000000, 2)) %>% 
  #rename columns
  #rename(., Comp = Completeness, Cont = Contamination, Mag_size = Genome_Size) %>% 
  remove_rownames() %>%
  arrange(match(Genome, rev(tree$tip.label))) #sort MAGs according to phylogenetic tree

head(mags_details)
```

##### MAGs circular tree
```{r circular tree, warning=FALSE, message = FALSE, fig.width= 11, fig.height=8}
# Generate the phylum color heatmap
phylum_heatmap <- mags_details %>%
  left_join(mags_phylum_colour, by=join_by(Genome == Genome)) %>%
   arrange(match(Genome, tree$tip.label)) %>% 
    select(-Phylum.y) %>%
    mutate(Phylum = factor(Phylum.x, levels = unique(Phylum.x))) %>%
    select(-Phylum.x) %>%
    column_to_rownames(var = "Genome")

# baseline tree
circular_tree <- force.ultrametric(tree, method="extend") %>% # extend to ultrametric for visualisation
    ggtree(., layout="circular", size=0.1, open.angle = 45)

# Add phylum ring
circular_tree <- gheatmap(circular_tree, heatmap, offset=0.60, width=0.1, colnames=FALSE) +
        geom_tiplab2(size=0.8, hjust=-0.1) +
        theme(legend.position = "right", legend.title = element_text("Phylum"), plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0))


# Flush color scale to enable a new color scheme in the next ring
circular_tree <- circular_tree + new_scale_fill()

# Add completeness ring
circular_tree <- circular_tree +
        new_scale_fill() +
        scale_fill_gradient(low = "#d1f4ba", high = "#f4baba") +
        geom_fruit(
                data=mags_details,
                geom=geom_bar,
                mapping = aes(x=Completeness, y=Genome, fill=Contamination),
                offset = 0.8,
                orientation="y",
              stat="identity")

# Add genome-size ring
circular_tree <-  circular_tree +
        new_scale_fill() +
        scale_fill_manual(values = "#cccccc") +
        geom_fruit(
             data=mags_details,
             geom=geom_bar,
             mapping = aes(x=mag_size, y=Genome),
                 offset = 0.05,
                 orientation="y",
         stat="identity")
# Add text
circle_tree <- circular_tree +
        annotate('text', x=1.7, y=0.0, label='             Phylum', size=3) +
        annotate('text', x=1.9, y=0.0, label='                         Genome quality', size=3) +
        annotate('text', x=2.1, y=0.0, label='                       Genome size', size=3) 

circle_tree %>% 
   open_tree(20) %>% rotate_tree(90)

```


#### Create composite figure

```{r composite fig, warning=FALSE, fig.width= 11, fig.height=9}
# First create a biplot chart
mag_stats_biplot <- mags_details %>%
        ggplot(aes(x=Completeness,y=Contamination,size=Genome_Size,color=Phylum)) +
              geom_point(alpha=0.7) +
                    ylim(c(10,0)) +
                    #scale_color_manual(values=colors_alphabetic) +
                labs(y= "Contamination", x = "Completeness") +
                    theme_classic() +
                  theme(legend.position = "bottom")

# X and Y axis boxplots to complement the plot and MAG quality statistics.

#contamination
mag_stats_cont <- mags_details %>%
            ggplot(aes(y=Contamination)) +
                    ylim(c(10,0)) +
                    geom_boxplot(colour = "#999999", fill="#cccccc") +
                    theme_void() +
                    theme(legend.position = "none",
                        axis.title.x = element_blank(),
                        axis.title.y = element_blank(),
                        axis.text.y=element_blank(),
                  axis.ticks.y=element_blank(),
                        axis.text.x=element_blank(),
                  axis.ticks.x=element_blank(),
                        plot.margin = unit(c(0, 0, 0.40, 0),"inches")) #add bottom-margin (top, right, bottom, left)

mag_stats_comp <-mags_details %>%
        ggplot(aes(x=Completeness)) +
                xlim(c(50,100)) +
                geom_boxplot(colour = "#999999", fill="#cccccc") +
                theme_void() +
                theme(legend.position = "none",
                    axis.title.x = element_blank(),
                    axis.title.y = element_blank(),
                    axis.text.y=element_blank(),
              axis.ticks.y=element_blank(),
                    axis.text.x=element_blank(),
              axis.ticks.x=element_blank(),
                    plot.margin = unit(c(0, 0, 0, 0.50),"inches")) #add left-margin (top, right, bottom, left)
```



```{r fig.width= 11, fig.height=8}
grid.arrange(grobs = list(mag_stats_comp,mag_stats_biplot,mag_stats_cont),
        layout_matrix = rbind(c(1,1,1,1,1,1,1,1,1,1,1,4),
                                                    c(2,2,2,2,2,2,2,2,2,2,2,3),
                                                    c(2,2,2,2,2,2,2,2,2,2,2,3),
                                                    c(2,2,2,2,2,2,2,2,2,2,2,3),
                                                    c(2,2,2,2,2,2,2,2,2,2,2,3),
                                                    c(2,2,2,2,2,2,2,2,2,2,2,3),
                                                    c(2,2,2,2,2,2,2,2,2,2,2,3),
                                                    c(2,2,2,2,2,2,2,2,2,2,2,3),
                                                    c(2,2,2,2,2,2,2,2,2,2,2,3),
                                                    c(2,2,2,2,2,2,2,2,2,2,2,3),
                                                    c(2,2,2,2,2,2,2,2,2,2,2,3),
                                                    c(2,2,2,2,2,2,2,2,2,2,2,3)))
```

MAGs are ordinated according to their completeness and contamination scores, while dot sizes indicate their genome-size(MBs).


```{r bad mags}
#Get list of present MAGs
present_MAGs <- count_table_cov_size %>%
        filter(rowSums(.) != 0) %>%
        rownames()

present_phylum <- unique(count[count$Genome %in% present_MAGs,]$Phylum)

cat("700 MAGs passed through a coverage filter of 0.03, resulting in ", length(present_MAGs), " remaining.
    \n\nOverall phyla detected: ", length(unique(count$Phylum)), "
    \nRemaining phyla: ", length(present_phylum),"\n")

cat("Bad MAGs with zero genome count for every samples:\n ")

bad.mags <- count_table_cov_size %>%
        filter(rowSums(.) == 0) %>%
        rownames()

#cat("Genome count of bad MAGs:\n ")

mags_details %>% filter(Genome %in% bad.mags) 

```

##### Heatmap for overall
```{r}
phylum_colour <- data.frame(Phylum = (unique(count_table$Phylum)), color = (hue_pal()(12)))
Phylum_colour <- setNames(phylum_colour$color, phylum_colour$Phylum)


logged <- count_table_cov_size[rownames(count_table_cov_size) != bad.mags, ]
mags_details.clm <- mags_details %>%
  column_to_rownames("Genome") %>% 
  filter(!(rownames(.) %in% bad.mags)) %>% 
  rownames_to_column("Genome")

colors <- rev(viridis::inferno(12))


annot.row <- data.frame(row.names = count_table$Genome, Phylum = as.factor(count_table$Phylum))
#phylum_colour <- phylum_colour[match(unique(phylum_colour$Phylum), unique(annot.row$Phylum)),]

anno.col <- data.frame(row.names = rownames(metadata), Emission = as.factor(metadata[,5]),Timepoint = as.factor(metadata[,2]), Breed = as.factor(metadata[,4]), Split = as.factor(metadata[,6]))



breakList = seq(0, 3, by = 0.25)

logged[is.na(logged)] <- 0

pheatmap(log2(logged + 1),
         cluster_rows = T,
         cluster_cols = F,
         color = colors,
         cellwidth = 10,
         breaks = breakList,
         annotation_col = anno.col,
         annotation_row = annot.row,
         annotation_colors = list(Phylum = Phylum_colour),
         show_rownames = F,
         legend = T,
         scale = "row",
         correlation = "Ward.D2"
         )
```
```{r}
metadata <- metadata[order(metadata$vsplit_old),]

pheatmap(log2(logged[,rownames(metadata)] + 1),
         cluster_rows = T,
         cluster_cols = F,
         color = colors,
         cellwidth = 10,
         breaks = breakList,
         annotation_col = anno.col,
         annotation_row = annot.row,
         annotation_colors = list(Phylum = Phylum_colour),
         show_rownames = F,
         legend = T,
         scale = "row",
         correlation = "Ward.D2"
         )
```
```{r}

metadata <- metadata[order(metadata$Emission),]

pheatmap(log2(logged[,rownames(metadata)] + 1),
         cluster_rows = T,
         cluster_cols = F,
         color = colors,
         cellwidth = 10,
         breaks = breakList,
         annotation_col = anno.col,
         annotation_row = annot.row,
         annotation_colors = list(Phylum = Phylum_colour),
         show_rownames = F,
         legend = T,
         scale = "row",
         correlation = "Ward.D2"
         )
```

```{r}
write.table(count_table_cov_size, file = "./Relative_abundance_0.3_thres.txt", quote = F, sep = "\t")

save(raw_count_table_cov_size, 
     bad.mags,
     tree,
     metadata,
     count_table,
     count_table_cov, # reads count table after 0.3
     count_table_cov_size, # genome counts table
     count_table_cov_size_pivot, 
     metadata, 
     mags_phylum_genomes_sorted, 
     mags_details, 
     file = "./01_Filter_count.RData")

#load("C:/Users/awala/OneDrive - Norwegian University of Life Sciences/Documents/Small_Cow/MAG/01_Filter_count.RData")
```



