---
title: "03_Diversity_analysis"
author: "Wanxin_Lai"
date: "2024-04-03"
output:
  rmarkdown::html_document:
    theme: lumen
    highlight: zenburn
    df_print: default
    toc: yes
    toc_float: TRUE
    keep_md: no
    code_folding: hide
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
setwd("C:/Users/awala/OneDrive - Norwegian University of Life Sciences/Documents/Small_Cow2_metaG/MAG/")
load("03_da.RData")
load("./02_differential_abundances.RData")

rm(emis_da, emis_da.genus, bias_correct_log_table,bias_correct_log_table_pivot)
```

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(DT)
library(distillR)
library(hilldiv2)
library(purrr)
library(broom.mixed)
library(ggtree)
library(phytools)
library(vegan)
library(ggpubr)
library(cowplot)
library(scales)
library(ggfortify)
library(ggforce)
library(data.table)
library(ggrepel)
```

# 1. Metabolic Capacity

#### Transforms raw functional annotation data into basal quantitative genome-inferred functional traits
```{r eval=FALSE, warning=FALSE, message=FALSE}
# annotations generated by DRAM
genome_annotations <- read_tsv("~/Small_Cow/MAG/annotations.tsv") %>%
    dplyr::rename(gene=1, genome=2, contig=3)
#write.csv(GIFT_db, file="gift_add.csv")

write.csv(genome_annotations, file = "Genome_annotations")

gift_cm <- read_csv("~/Small_Cow/MAG/gift_add.csv")
gift_cm <- gift_cm %>% select(-1)

genome_gifts <- distill(genome_annotations, gift_cm , genomecol=2,annotcol=c(9,10,19))

#save(genome_gifts, gift_cm, file = "03_da.RData")
rm(genome_annotations)
```

Relying on the hierarchical structure of the distillR database, basal GIFTs can be aggregated into larger clusters at the compound (170 compounds), function (10 functions) and domain (3 domains) level.

```{r}
#Aggregate bundle-level GIFTs into the compound level
GIFTs_elements <- to.elements(genome_gifts, gift_cm)

GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% rownames(count_table_cov_size %>% 
  select(-c("D55T5", "D67T5"))), ] %>% 
  as.data.frame() %>% 
  select_if(~ sum(.) != 0) #remove column with only 0s

```

```{r}
#Aggregate element-level GIFTs into the function level
GIFTs_functions <- to.functions(GIFTs_elements,gift_cm)

GIFTs_functions_filtered <- GIFTs_functions[rownames(GIFTs_functions) %in% rownames(count_table_cov_size %>% 
  select(-c("D55T5", "D67T5"))), ] %>% 
  as.data.frame() %>% 
  select_if(~ sum(.) != 0)

#Aggregate function-level GIFTs into overall Biosynthesis, Degradation and Structural GIFTs
GIFTs_domains <- to.domains(GIFTs_functions,gift_cm)

GIFTs_domains_filtered <- GIFTs_domains[rownames(GIFTs_domains) %in% rownames(count_table_cov_size %>% 
  select(-c("D55T5", "D67T5"))), ] %>% 
  as.data.frame() %>% 
  select_if(~ sum(.) != 0)

# Separative approach (considering relative abundances)
# Get community-weighed average GIFTs per sample
GIFTs_elements_community <- to.community(GIFTs_elements_filtered, (count_table_cov_size %>% 
  select(-c("D55T5", "D67T5"))) %>%
    tss(), gift_cm)
GIFTs_functions_community <- to.community(GIFTs_functions_filtered, (count_table_cov_size %>% 
  select(-c("D55T5", "D67T5"))) %>%
    tss(), gift_cm)
GIFTs_domains_community <- to.community(GIFTs_domains_filtered, (count_table_cov_size %>% 
  select(-c("D55T5", "D67T5"))) %>%
    tss(), gift_cm)
```

### Matabolic capacity with tSNE
```{r}
#set.seed(990)
#perplexity numeric; Perplexity parameter (should not be bigger than 3 * perplexity < nrow(X) - 1, see details for interpretation)
# reverse-calculate the highest acceptable perplexity:
ordination <- Rtsne::Rtsne(X=GIFTs_elements, dims = 2, perplexity = floor((nrow(GIFTs_elements) - 1) / 3), check_duplicates = FALSE)

GIFTs_elements_MCI <- GIFTs_elements %>%
  as.data.frame %>% 
  mutate(Avg.MCI=rowMeans(data.frame(GIFTs_elements)))

# Plot the ordination
element_ordination <- ordination$Y %>%
  as.data.frame() %>%
  mutate(Genome=rownames(GIFTs_elements)) %>%
  dplyr::rename("tSNE1"="V1", "tSNE2"="V2") %>%
  left_join(mags_details, by = join_by(Genome == Genome)) %>%
  ggplot(aes(x = tSNE1, y = tSNE2, color=Phylum, group = Phylum, label = Phylum)) +
  scale_fill_manual(values = phylum_colour) +
    geom_point(size = 4) +
    theme_classic() +
  theme(legend.key.size = unit(0.5, 'cm'), #change legend key size
        legend.key.height = unit(0.1, 'cm'), #change legend key height
        legend.key.width = unit(0.1, 'cm'), #change legend key width
        legend.title = element_text(size=10), #change legend title font size
        legend.text = element_text(size=10)) #change legend text font size 

# Add grouping variable
circle <- GIFTs_elements_MCI %>%
  as.data.frame() %>% 
  mutate(group = ifelse(Avg.MCI > 0.33, "HighMC", ifelse(Avg.MCI < 0.030 & Avg.MCI > 0.001, "LowMC", NA)))

# DistillR MCI value
element_ordination_MCI <- ordination$Y %>%
  as.data.frame() %>%
  mutate(Genome=rownames(GIFTs_elements_MCI)) %>%
  dplyr::rename("tSNE1" = "V1", "tSNE2"="V2") %>% 
  left_join(mags_details, by = join_by(Genome == Genome)) %>%
  merge(data.frame(Avg.MCI = circle[,"Avg.MCI"],group = circle[,"group"], row.names = rownames(circle)), by.x = "Genome", by.y = 0) %>% 
  ggplot(aes(x = tSNE1, y = tSNE2, color=Avg.MCI, group = group)) + geom_point(size = 2, shape = 16, alpha = 0.8) +
  # Filter data for HighMC and LowMC groups before applying stat_ellipse
  #geom_ellipse(data = . %>% filter(group == "HighMC"),
      #         aes(alpha =0, color = "blue")) +
 geom_mark_ellipse(data = . %>% filter(group == "HighMC")) +
  geom_mark_ellipse(data = . %>% filter(group == "LowMC")) +
  scale_colour_gradientn(colours = rev(terrain.colors(10))) +
  theme_classic() 

plot_grid(element_ordination, element_ordination_MCI, nrow = 3)
  
```
Highest vs lowest MCI MAGs

```{r, eval = FALSE}
## need 03_hmsc output

breed.hmsc_ord <- ordination$Y %>%
  as.data.frame() %>%
  mutate(Genome=rownames(GIFTs_elements)) %>%
  dplyr::rename("tSNE1"="V1", "tSNE2"="V2") %>% 
  left_join(post_beta %>% rownames_to_column("Genome") %>% select(Genome, Emissionlow, Vsplitgroup2, BreedLuing), by = join_by(Genome == Genome)) %>% 
  
  ggplot(aes(x = tSNE1, y = tSNE2, colour=Emissionlow)) + 
  geom_point(size = 4) +
  theme_classic() +
  scale_color_manual(values = c("#b2b530","grey","#be3e2b"))+
  theme(legend.key.size = unit(0.5, 'cm'), #change legend key size
        legend.key.height = unit(0.1, 'cm'), #change legend key height
        legend.key.width = unit(0.1, 'cm'), #change legend key width
        legend.title = element_text(size=10), #change legend title font size
        legend.text = element_text(size=10)) #change legend text font size 

```

```{r}
plot_grid(element_ordination, element_ordination_MCI,breed.hmsc_ord, nrow = 3)
```

```{r}
high.MAGs <- read.csv("~/Small_Cow2_metaG/MAG/hmsc_split_seaweed/modelsminus2/panels/high_emi_MAGs.csv" , row.names = 1) %>% 
   merge(., (GIFTs_elements_MCI %>%
          rownames_to_column("MAG") %>%
          select(MAG, Avg.MCI)), by.x = "genome", by.y = "MAG")

low.MAGs <- read.csv("~/Small_Cow2_metaG/MAG/hmsc_split_seaweed/modelsminus2/panels/low_emi_MAGs.csv") %>% 
  merge(., (GIFTs_elements_MCI %>%
          rownames_to_column("MAG") %>%
          select(MAG, Avg.MCI)), by.x = "genome", by.y = "MAG")

t.result <- t.test(high.MAGs$Avg.MCI, low.MAGs$Avg.MCI)

print(t.result)
```

```{r}
wilcox.test(high.MAGs$Avg.MCI, low.MAGs$Avg.MCI)
```

```{r}

#perplexity numeric; Perplexity parameter (should not be bigger than 3 * perplexity < nrow(X) - 1, see details for interpretation)
# reverse-calculate the highest acceptable perplexity:
ordination <- Rtsne::Rtsne(X=GIFTs_functions, dims = 2, perplexity = floor((nrow(GIFTs_functions) - 1) / 3), check_duplicates = FALSE)

# Plot the ordination
functions_ordination <- ordination$Y %>%
  as.data.frame() %>%
  mutate(Genome=rownames(GIFTs_elements)) %>%
  #mutate(Sample = GIFTs_elements$Sample)
  dplyr::rename("tSNE1"="V1", "tSNE2"="V2") %>% 
  left_join(mags_details, by = join_by(Genome == Genome)) %>%
  ggplot(aes(x = tSNE1, y = tSNE2, color=Phylum, group = Phylum, label = Phylum)) +
  scale_fill_manual(values = phylum_colour) +
    geom_point(size = 4) +
    #geom_path(aes(linetype=group),color="#999999", alpha=0.5, arrow = arrow(length=unit(0.30,"cm"), ends="last", type = "closed")) +
    theme_classic()

```

```{r}
metadata$Emission <- factor(metadata$Emission)

raw.pca <- prcomp(GIFTs_functions)

summary(raw.pca)

pca_pl <- raw.pca$x %>%
  as.data.frame() %>%
  merge(mags_details, by.x = 0, by.y = "Genome") %>% 
  ggplot(aes(x = PC1, y = PC2, color=Phylum, group = Phylum, label = Phylum)) +
  scale_fill_manual(values = phylum_colour) +
    geom_point(size = 4) +
    theme_classic() +
  theme(legend.key.size = unit(0.5, 'cm'), #change legend key size
        legend.key.height = unit(0.1, 'cm'), #change legend key height
        legend.key.width = unit(0.1, 'cm'), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=5)) #change legend text font size

```



```{r}
pca_mci_pl <- raw.pca$x %>%
  as.data.frame() %>%
  merge(mags_details, by.x = 0, by.y = "Genome") %>% 
  merge(data.frame(Avg.MCI = GIFTs_elements_MCI[,"Avg.MCI"], row.names = rownames(GIFTs_elements_MCI)), by.x = "Row.names", by.y = 0) %>% 

ggplot(aes(x = PC1, y = PC2, color=Avg.MCI)) + geom_point(size = 2, shape = 16, alpha = 0.8) +
  scale_colour_gradientn(colours = rev(terrain.colors(10))) +
  theme_classic() 


plot_grid(pca_pl, pca_mci_pl, nrow = 2)
```


## Patescibacteria(Lowest MCI)
```{r}
datatable(raw.pca$x %>%
  as.data.frame() %>%
  merge(mags_details, by.x = 0, by.y = "Genome") %>%  
  filter(., Phylum  == "Patescibacteria") %>% 
  select(., !contains("PC"))
)
```



```{r warning=F}
width_scale <- 12

dftxt <- raw.pca$x %>%
  as.data.frame() %>%
  merge(mags_details, by.x = 0, by.y = "Genome") %>%  
  merge(circle %>% 
          rownames_to_column("Genome") %>% 
          select("Genome", "Avg.MCI", "group"), by.x = "Row.names", by.y = "Genome") %>% 
  filter(., group  == "LowMC") %>%
  mutate(zoom = TRUE )   

raw.pca$x %>%
  as.data.frame() %>%
  merge(mags_details, by.x = 0, by.y = "Genome") %>% 
  ggplot(aes(x = PC1, y = PC2, color=Phylum, group = Phylum, label = Genus)) +
  scale_fill_manual(values = phylum_colour) +
    geom_point(size = 4) +
    theme_classic() +
  guides(fill = guide_legend(override.aes = list(size = 4), ncol = 1))+
  facet_zoom(xlim = c(-1.1, -1.6), zoom.data=.[[dftxt$zoom]]) +
  geom_text_repel(data = dftxt, aes(label = Genus), max.overlaps = 5, force = 50, show.legend = F) + # removed from the main plot
  theme(legend.position = "none",
        legend.title = element_text(size=5,face="bold"), #change legend title font size
        legend.text = element_text(size=5),
        legend.key.size = grid::unit(width_scale/60, "cm"),
                   legend.key.width = grid::unit(width_scale/60, "cm"))#change legend text font size

```


## Actinobacteriota (Highest MCI)
```{r}
acino <- GIFTs_elements_MCI %>%
  as.data.frame()%>%
  top_n(2,Avg.MCI) %>%
  rownames(.)

datatable(raw.pca$x %>%
  as.data.frame() %>%
  merge(mags_details, by.x = 0, by.y = "Genome") %>%  
  filter(Row.names  %in% acino) %>% 
  select(., !contains("PC"))
)
```

### Actinobacter(Highest MCI)
```{r}

width_scale <- 12

dftxt <- raw.pca$x %>%
  as.data.frame() %>%
  merge(mags_details, by.x = 0, by.y = "Genome") %>%  
filter(Row.names  %in% acino) %>%
  mutate(zoom = TRUE )   

raw.pca$x %>%
  as.data.frame() %>%
  merge(mags_details, by.x = 0, by.y = "Genome") %>% 
  ggplot(aes(x = PC1, y = PC2, color=Phylum, group = Phylum, label = Genus)) +
  scale_fill_manual(values = phylum_colour) +
    geom_point(size = 4) +
    theme_classic() +
  guides(fill = guide_legend(override.aes = list(size = 4), ncol = 1))+
  #facet_zoom(xlim = c(0.65, 0.9), zoom.data=.[[dftxt$zoom]]) +
  geom_text_repel(data = dftxt, aes(label = Genus), max.overlaps = 30, force = 20, show.legend = F) + # removed from the main plot
  theme(legend.position = "none",
        legend.title = element_text(size=5,face="bold"), #change legend title font size
        legend.text = element_text(size=5),
        legend.key.size = grid::unit(width_scale/60, "cm"),
                   legend.key.width = grid::unit(width_scale/60, "cm"))#change legend text font size
```

```{r}

# set a cut off to filter important for loadings
loading_cutoff <- sqrt(1/ncol(GIFTs_functions))

# Create an empty list for those that pass the cutoff
passing_rownames <- vector()

# setting keep_row = FALSE
for(pc_row in 1:nrow(raw.pca$rotation)) {
  keep_row <- FALSE
  
  # for the first 1 PCs
  for (pc_col in 1:2) {
    if (abs(raw.pca$rotation[pc_row, pc_col]) > loading_cutoff) {
      keep_row <- TRUE
    }
  }
  
  if (keep_row == TRUE) {
    passing_rownames <- c(passing_rownames, rownames(raw.pca$rotation)[pc_row])
  }
}

# now we know which genes have the most influence on each selected PC axis.
cat(paste(length(passing_rownames), " features pass the loading cutoff of ", round(loading_cutoff, 4), sep = "" ))

#select non duplicated code function
cf <- gift_cm %>% distinct(gift_cm[,3], .keep_all=T)

data.frame(cf)[,-7] %>% filter(Code_function %in% passing_rownames)

```


```{r Function}
# Function to replace column name
replace_column_name <- function(input_df, reference_db, index_input_column, index_db_column) {
  
  #input_df <- GIFTs_functions_community
  #reference_db <- gift_cm
  #index_db_column <- 5
  #index_input_column <- 3
  
  
  input_df <- data.frame(input_df)
  
  custom_order <- c(colnames(input_df))
  
  unique_row_df <- reference_db %>% distinct(reference_db[,index_input_column], .keep_all=T) %>% 
    filter(.[[index_input_column]] %in% custom_order) %>% 
    relocate(any_of(custom_order))
  
  #Reorder the dataframe and print only the column of interest
  #new_colname <- unique_row_df[, c(index_input_column,index_db_column)]
  
  new_colname <- unique_row_df %>% mutate(., x = paste(.[[index_input_column]], .[[index_db_column]])) %>% select(x)

  for(i in 1:ncol(input_df)){
    colnames(input_df)[i] <- new_colname$x[i]
  }
  return(input_df)
  
}

```


### Time series community functions
```{r}
GIFTs_functions_names <- replace_column_name(GIFTs_functions_community, gift_cm, 3, 5)

GIFTs_functions_names %>%
  # z standardisation
  mutate_all(~scale(.) %>% as.vector) %>%
    as.data.frame() %>%
    rownames_to_column(var="Sample") %>%
    pivot_longer(!Sample,names_to="trait",values_to="GIFT_sd") %>%
    merge(metadata, by.x = "Sample", by.y = 0) %>%
    ggplot(aes(x=trait,y=Sample,group = Time, fill=GIFT_sd)) +
        geom_tile(colour="white", size=0.2)+
        scale_fill_gradientn(colours=rev(c("#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#e6f598", "#abdda4", "darkgreen")))+
  theme(legend.position = "right", axis.text.x = element_text(angle = 40, vjust = 0.5, hjust=1, margin = margin(t=-40), size = 7)) +

        facet_grid(Time ~., scales="free",space="free")
```


#### Emission communiy functions
```{r}
GIFTs_functions_names %>%
  mutate_all(~scale(.) %>% as.vector) %>% 
    as.data.frame() %>%
    rownames_to_column(var="Sample") %>%
    pivot_longer(!Sample,names_to="trait",values_to="GIFT_sd") %>%
    merge(metadata, by.x = "Sample", by.y = 0) %>%
    ggplot(aes(x=trait,y=Sample,group = Emission, fill=GIFT_sd)) +
        geom_tile(colour="white", size=0.01)+
        scale_fill_gradientn(colours=rev(c("#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#e6f598", "#abdda4", "darkgreen")))+
  theme(legend.position = "right", axis.title.x = element_blank(), axis.text.x = element_text(angle = 30, vjust = 0.5, hjust=1, margin = margin(t=-33, b = 35, l = 10), size = 7)) +
        facet_grid(Emission ~., scales="free",space="free")
```

```{r}
write.csv(GIFTs_functions_names %>%
  mutate_all(~scale(.) %>% as.vector) %>% 
    as.data.frame() %>%
    rownames_to_column(var="Sample"), "./output_dist_R/GIFTs_functions_Emission.csv")
```

```{r}
ordination_samples <- Rtsne::Rtsne(X=GIFTs_functions_names, dims = 2, perplexity = floor((nrow(GIFTs_functions_names) - 1) / 3), check_duplicates = FALSE)

# Plot the ordination
functions_samples_ordination <- ordination_samples$Y %>%
  as.data.frame() %>%
  mutate(Sample=rownames(GIFTs_functions_names)) 

#functions_samples_ordination$Emission <- as.factor(GIFTs_functions_names$Emission)

functions_samples_ordination %>% 
  rename(tSNE1="V1", tSNE2="V2") %>%
 merge(metadata, by.x = "Sample", by.y = 0) %>%
  ggplot(aes(x = tSNE1, y = tSNE2, shape=vsplit_old, color = Emission, label = Time)) +
  #scale_fill_manual(values = phylum_colour) +
    geom_point(size = 4) +
  geom_text(aes(label=as.character(Animal)),hjust=-0.3,vjust=0.1)+
    #geom_path(aes(linetype=group),color="#999999", alpha=0.5, arrow = arrow(length=unit(0.30,"cm"), ends="last", type = "closed")) +
    theme_classic()

```

##### Significance test on community functions

```{r}
functions_test <- GIFTs_functions_names %>%
  as.data.frame() %>%
  rownames_to_column("Sample") %>%
  pivot_longer(!Sample, names_to = "trait", values_to = "GIFT") %>%
  merge(metadata, by.x = "Sample", by.y = 0) %>%
  group_by(trait) %>%
  mutate(model_result = list(lmerTest::lmer(GIFT ~ Emission + (1 | Animal)))) %>%
  ungroup() %>%
  select(trait,model_result) %>%
  unique() %>%
  mutate(p_value = map_dbl(model_result, ~broom.mixed::tidy(.) %>% filter(term == "Emissionlow") %>% pull(p.value))) %>%
  mutate(p_value_adj = p.adjust(p_value, method = "fdr")) %>%
  select(trait, p_value, p_value_adj) %>% 
  arrange(-p_value)

datatable(functions_test)
```

```{r}

colors <- rev(viridis::inferno(12))

#select non duplicated code function
cf <- gift_cm %>% distinct(gift_cm[,3], .keep_all=T)

metadata <- metadata[order(metadata$vsplit_old),]

anno.col <- data.frame(row.names = rownames(metadata), Emission = as.factor(metadata[,5]),Timepoint = as.factor(metadata[,2]), Split = as.factor(metadata[,6]))

heatmp <- GIFTs_functions_names %>%
  as.data.frame() %>% 
  # z standardisation
  mutate_all(~scale(.) %>% as.vector)


pheatmap::pheatmap(t(heatmp),
                   cluster_rows = T,
                   cluster_cols = F,
                   color = colors,
                   cellwidth = 10,
                   #breaks = breakList,
                   annotation_col = anno.col,
                   #annotation_row = annot.row,
                   show_rownames = T,
                   legend = T,
                   scale = "row",
                   correlation = "Ward.D2",
                   )
```

```{r}
metadata <- metadata[order(metadata$Emission),]

pheatmap::pheatmap(t(heatmp[rownames(metadata),]),
                   cluster_rows = T,
                   cluster_cols = F,
                   color = colors,
                   cellwidth = 10,
                   #breaks = breakList,
                   annotation_col = anno.col,
                   #annotation_row = annot.row,
                   show_rownames = T,
                   legend = T,
                   scale = "row",
                   correlation = "Ward.D2",
                   )
         

```



#### Timeseries community elements
```{r fig.height=10, fig.width=14}
GIFTs_elements_names <- replace_column_name(GIFTs_elements_community, gift_cm, 2, 6)

GIFTs_elements_names %>%
  mutate_all(~scale(.) %>% as.vector) %>%
    as.data.frame() %>%
    rownames_to_column("Sample") %>%
    pivot_longer(!Sample,names_to="trait",values_to="GIFT_sd") %>%
    merge(metadata, by.x = "Sample", by.y = 0) %>%
    ggplot(aes(x=trait,y=Sample,group = Time, fill=GIFT_sd)) +
        geom_tile(colour="white", size=0.3)+
        scale_fill_gradientn(colours=rev(c("#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#e6f598", "#abdda4", "darkgreen")))+
  theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, margin = margin(t=-0), size = 7)) +

        facet_grid(Time ~., scales="free",space="free")
```

```{r}
write.csv(GIFTs_elements_names %>%
  mutate_all(~scale(.) %>% as.vector) %>%
    as.data.frame() %>%
    rownames_to_column("Sample"), "./output_dist_R/GIFTs_compounds_Emission.csv") 
```


#### Emission community elements
```{r fig.height=6, fig.width=12}

GIFTs_elements_names %>%
  mutate_all(~scale(.) %>% as.vector) %>%
    as.data.frame() %>%
    rownames_to_column("Sample") %>%
    pivot_longer(!Sample,names_to="trait",values_to="GIFT_sd") %>%
    merge(metadata, by.x = "Sample", by.y = 0) %>%
    ggplot(aes(x=trait,y=Sample,group = Time, fill=GIFT_sd)) +
        geom_tile(colour="white", size=0.2)+
        scale_fill_gradientn(colours=rev(c("#d53e4f", "#f46d43", "#fdae61", "#fee08b", "#e6f598", "#abdda4", "darkgreen")))+
  theme(legend.position = "right", axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, margin = margin(t=-0), size = 7)) +

        facet_grid(Emission ~., scales="free",space="free")

```

##### Significance test on elements community
```{r}
element_test <- GIFTs_elements_names %>%
  as.data.frame() %>%
    rownames_to_column("Sample") %>%
    pivot_longer(!Sample,names_to="trait",values_to="GIFT") %>%
    merge(metadata, by.x = "Sample", by.y = 0) %>%
  group_by(trait) %>%
  mutate(model_result = list(lmerTest::lmer(GIFT ~ Emission + (1 | Animal)))) %>%
  ungroup() %>%
  select(trait,model_result) %>%
  unique() %>%
  mutate(p_value = map_dbl(model_result, ~broom.mixed::tidy(.) %>% filter(term == "Emissionlow") %>% pull(p.value))) %>%
  mutate(p_value_adj = p.adjust(p_value, method = "fdr")) %>%
  select(trait, p_value, p_value_adj) %>% 
  arrange(-(p_value))

datatable(element_test)
```

The following convention for symbols indicating statistical significance:

ns: p > 0.05

*: p <= 0.05

**: p <= 0.01

***: p <= 0.001

****: p <= 0.0001

## Diversity analysis

```{r}
#Filter badMAGs
good.mags <- count_table_cov_size %>% 
  select(-c("D55T5", "D67T5"))%>%
  filter(rowSums(.) != 0) %>%
        rownames()

good.mags.count <- count_table_cov %>% 
  #select(-c("D55T5", "D67T5"))%>%
  filter(rowSums(.) != 0) %>%
        rownames()

tree <- read.tree("./results_ac2/mashtree/mashtree.newick")

genome_tree <- force.ultrametric(tree,method="extend")


```

Hill numbers gives the number of equally abundant OTUs that would be needed to give the same value of diversity.

```{r}

new_cov <- count_table_cov_size %>% 
 select(-c("D55T5", "D67T5"))

# Calculate Hill numbers
richness <- new_cov %>% 
  select(where(~ !all(. == 0))) %>% #select columns where not all values in that column are equal to zero.
  hilldiv2::hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "Sample")

# Neutral Hill numbers only rely on the relative abundance information to compute diversity
neutral <-new_cov %>%
  select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "Sample")

# Phylogenetic Hill numbers also incorporate branch-length information of the phylogenetic tree of the MAGs
phylogenetic <- new_cov %>%
  select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, tree = genome_tree) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(phylogenetic = 1) %>%
  rownames_to_column(var = "Sample")

# Aggregate basal GIFT into elements
dist <- genome_gifts %>%
  to.elements(., gift_cm) %>%
  traits2dist(., method = "gower")
```


```{r}
dist_tibb <- genome_gifts[good.mags,] %>%
  to.elements(., gift_cm) %>%
  traits2dist(., method = "gower")
  
genome_count <- as_tibble(new_cov[rownames(dist_tibb),] %>% rownames_to_column(var = "Genome"))

# the functional Hill numbers account for the functional differences between MAGs, based on a distance matrix derived from distilled functional traits of MAGs 
functional <- genome_count %>%
  column_to_rownames("Genome") %>% 
  #select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, dist = dist_tibb) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(functional = 1) %>%
  rownames_to_column(var = "Sample") %>%
  mutate(functional = if_else(is.nan(functional), 1, functional))

```


```{r}
# Merge all metrics
alpha_div <- richness %>%
 # full_join(neutral, by = join_by(Sample == Sample)) %>%
  full_join(phylogenetic, by = join_by(Sample == Sample)) %>%
  full_join(functional, by = join_by(Sample == Sample))
```

#### Compare alpa diversity across time


```{r fig.width=10}
alpha_div_plo <- alpha_div %>%
  pivot_longer(-Sample, names_to = "metric", values_to = "Hill") %>%
  merge(metadata, by.x = "Sample", by.y = 0) %>%
  mutate(metric=factor(metric,levels=c("richness","phylogenetic","functional")))

alpha_div_plo %>%
      ggplot(aes(y = Hill, x = Animal, group=Animal, color=Breed, fill=Breed)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(alpha=0.5) +
      scale_color_manual(values=c("#e5bd5b", "#6b7398","#e2815a", "#876b96")) +
      scale_fill_manual(values=c("#e5bd5b50", "#6b739850","#e2815a50", "#876b9650")) +
      facet_wrap(. ~ metric, scales = "free", ncol=4) +
      coord_cartesian(xlim = c(1, NA)) +
      theme_classic() +
      stat_compare_means() +
      theme(
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text())

```

No differences exist between breed.
```{r fig.width=10}
alpha_div_plo <- alpha_div %>%
  pivot_longer(-Sample, names_to = "metric", values_to = "Hill") %>%
  merge(metadata, by.x = "Sample", by.y = 0) %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic","functional"))) %>% 
  filter(Emission == "high")# %>% 
  #unite("Breed", c(Breed, Emission), remove = FALSE)

alpha_div_plo %>%
      ggplot(aes(y = Hill, x = Animal, group=Animal, color=Animal, fill=Breed)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(alpha=0.5) +
      scale_color_manual(values=c("#e5bd5b", "#6b7398","#e2815d", "#876b96")) +
      scale_fill_manual(values=c("#e5bd5b30", "#6b739850","#e2815a90", "#876b9650")) +
      facet_wrap(. ~ metric, scales = "free", ncol=4) +
      coord_cartesian(xlim = c(1, NA)) +
      theme_classic() +
      stat_compare_means(aes(label = paste("          p = ", ..p.format.., sep = ""))) +
  xlab("Animals (High CH4 emitters)")+
      theme(
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .4, color = "grey"),
        #axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(size = 10),
        strip.text = element_text(size = 12))
#"#e2815d"
```

```{r}

```

Functional difference between Angus high and Angus low(high and low methane emission are functionally different in different breed...)

```{r fig.width=10}
alpha_div %>%
  pivot_longer(-Sample, names_to = "metric", values_to = "Hill") %>%
  merge(metadata, by.x = "Sample", by.y = 0) %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic","functional"))) %>% 
      ggplot(aes(y = Hill, x = Emission, group=Animal, color=Emission, fill=Emission)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(alpha=0.5) +
      scale_color_manual(values=c("#e5bd5b", "#6b7398")) +
      scale_fill_manual(values=c("#e5bd5b50", "#6b739850")) +
      facet_wrap(. ~ metric, scales = "free", ncol=4) +
      coord_cartesian(xlim = c(1, NA)) +
      stat_compare_means() +
      theme_classic() +
      theme(
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text())
```

No differences between emission group.
```{r fig.width=10}
alpha_div %>%
  pivot_longer(-Sample, names_to = "metric", values_to = "Hill") %>%
  merge(metadata, by.x = "Sample", by.y = 0) %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic","functional"))) %>% 
  unite("Emission", c(Time, Emission), remove = FALSE) %>% 
      ggplot(aes(y = Hill, x = Time, group=Emission, color=Emission, fill=Emission)) +
      geom_boxplot(outlier.shape = NA) +
      geom_point(alpha=0.5) +
      scale_color_manual(values=c("#EEB42290","#FFEC8B90","#6b739860","#B0C4DE80","#FFA07A","#e2815a90", "#9370DB90","#876b9650", "#EE6EB460", "#FFB6C150")) +
      scale_fill_manual(values=c("#EEB42290","#FFEC8B90","#6b739860","#B0C4DE80","#FFA07A","#e2815a90", "#9370DB90","#876b9650", "#EE6EB460", "#FFB6C150")) +
      facet_wrap(. ~ metric, scales = "free", ncol=4) +
      coord_cartesian(xlim = c(1, NA)) +
      theme_classic() +
  stat_compare_means(aes(label = ..p.format..)) +
      theme(
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text())

```

```{r fig.width=10}
alpha_div %>%
  pivot_longer(-Sample, names_to = "metric", values_to = "Hill") %>%
  merge(metadata, by.x = "Sample", by.y = 0) %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic","functional"))) %>% 
  unite("Emission", c(vsplit_old, Emission), remove = FALSE) %>% 
      ggplot(aes(y = Hill, x = vsplit_old, group=Animal, color=Emission, fill=Emission)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(alpha=0.5) +
      scale_color_manual(values=c("#EEB42290","#FFEC8B90","#6b739860","#B0C4DE80")) +
      scale_fill_manual(values=c("#EEB42290","#FFEC8B90","#6b739860","#B0C4DE80")) +
      facet_wrap(. ~ metric, scales = "free", ncol=4) +
      coord_cartesian(xlim = c(1, NA)) +
      theme_classic() +
   stat_compare_means(aes(label = ..p.format..)) +
      theme(
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text())
```

No significant different found between low and high emission in every split.
```{r fig.width=10}
alpha_div %>%
  pivot_longer(-Sample, names_to = "metric", values_to = "Hill") %>%
  merge(metadata, by.x = "Sample", by.y = 0) %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic","functional"))) %>% 
      ggplot(aes(y = Hill, x = vsplit_old, group=Animal, color=vsplit_old, fill=vsplit_old)) +
      geom_boxplot(outlier.shape = NA, alpha = 0.5) +
      geom_jitter(alpha=0.5) +
      #scale_color_manual(values=c("#EEB42290","#FFEC8B","#6b739860","#B0C4DE80","#FFA07A","#e2815a90", "#9370DB90","#876b9650", "#EE6EB460", "#FFB6C150")) +
      #scale_fill_manual(values=c("#EEB42290","#FFEC8B","#6b739860","#B0C4DE80","#FFA07A","#e2815a90", "#9370DB90","#876b9650", "#EE6EB460", "#FFB6C150")) +
      facet_wrap(. ~ metric, scales = "free", ncol=4) +
      coord_cartesian(xlim = c(1, NA)) +
      theme_classic() +
   stat_compare_means() +
      theme(
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank())
```



###Hill numbers pairwise dissimilarities
Using hillpair(), dissimilarities are computed for every pairwise combination of samples, rather than the entire set of sample.

```{r}
beta_q0n <- count_table_cov_size  %>%
  hillpair(., q = 0)

beta_q1n <- count_table_cov_size %>%
  hillpair(., q = 1)

beta_q1p <- count_table_cov_size %>%
  hillpair(., q = 1, tree = genome_tree)

beta_q1f <- genome_count %>%
  filter(Genome %in% labels(dist_tibb)[[1]]) %>%
  column_to_rownames("Genome") %>%
  hillpair(., q = 1, dist = dist_tibb)
```

#####V (Sørensen-type turnover): it quantifies the normalised species turnover rate in a sample relative to one sample (i.e., alpha diversity).

#####C (Sørensen-type overlap): it quantifies the effective average proportion of shared OTUs/ASVs/MAGs in samples. Therefore, this metric quantifies similarity from the perspective of a single sample.

#####S (Jaccard-type turnover-complement): it quantifies the complement of the normalised species turnover rate in a sample relative to the total pool of samples (i.e., gamma diversity).

##### To measure the compositional changes across different time points, as well as between all time points to measure overall variability within an individual through time. 
Values closer to 1 indicate high compositional differences, whereas values close to 0 indicate less compositional differences.
```{r}
beta_neutral_nmds <-  vegan::metaMDS(beta_q1n$S, trymax = 500, k = 2, verbosity = FALSE, trace=FALSE) %>%
  vegan::scores() 

beta_neutral_nmds_vsplit_plot <- beta_neutral_nmds %>%
  as_tibble(., rownames = "Sample") %>%
  merge(metadata, by.x = "Sample", by.y = 0) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color=Emission, shape=vsplit_old)) +
    geom_point(size = 4) +
  #geom_text(aes(label=as.character(Animal)),hjust=-0.3,vjust=0.1)+
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
    theme_classic() +
  xlab("No split contribution to beta diversities")+
  ylab("")
```

```{r}
beta_neutral_nmds <-  vegan::metaMDS(beta_q1f$S, trymax = 500, k = 2, verbosity = FALSE, trace=FALSE) %>%
  vegan::scores() 

beta_neutral_nmds_time_plot <- beta_neutral_nmds %>%
  as_tibble(., rownames = "Sample") %>%
  merge(metadata, by.x = "Sample", by.y = 0) %>%
  
   group_by(Time) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  #mutate(Animal=factor(Animal, levels=c("A02", "A67", "A51", "A28", "A55", "A01"))) %>%
  ungroup() %>%
  
  ggplot(aes(x = NMDS1, y = NMDS2, color=Time, shape=Emission, group = Time, label = Animal)) +
    geom_point(size = 4) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
 # geom_text(aes(label=as.character(Animal)),hjust=-0.3,vjust=0.1)+
  theme_classic() +
  xlab("Time contributes functional differences")+
  ylab("")
```
NMDS 1 show that time differences contributes to neutral diversities.

```{r}
beta_phyloge_nmds <-  vegan::metaMDS(beta_q1p$S, trymax = 500, k = 2, verbosity = FALSE, trace=FALSE) %>%
  vegan::scores() 

beta_phyloge_nmds_plot <- beta_phyloge_nmds %>%
  as_tibble(., rownames = "Sample") %>%
  merge(metadata, by.x = "Sample", by.y = 0) %>%
  
   group_by(Emission) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  #mutate(Animal=factor(Animal, levels=c("A02", "A67", "A51", "A28", "A55", "A01"))) %>%
  ungroup() %>%
  
  ggplot(aes(x = NMDS1, y = NMDS2, color=Emission, shape=Emission, group = Emission, label = Animal)) +
    geom_point(size = 4) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  #geom_text(aes(label=as.character(Animal)),hjust=-0.3,vjust=0.1)+
  theme_classic() +
  xlab("Emission category contributes to phylogeny differences")+
  ylab("")
```
The is a phylogeny and functinal differences in term of beta diversities between low and high emission group.
```{r}
beta_function_nmds <-  vegan::metaMDS(beta_q1f$S, trymax = 500, k = 2, verbosity = FALSE, trace=FALSE) %>%
  vegan::scores() 

beta_function_nmds_plot <- beta_function_nmds %>%
  as_tibble(., rownames = "Sample") %>%
  merge(metadata, by.x = "Sample", by.y = 0) %>%
  
   group_by(Breed) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  #mutate(Animal=factor(Emission, levels=c("A02", "A67", "A51", "A28", "A55", "A01"))) %>%
  ungroup() %>%
  
  ggplot(aes(x = NMDS1, y = NMDS2, color=Breed, shape=Breed, group = Breed, label = Animal)) +
    geom_point(size = 4) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9) +
  #geom_text(aes(label=as.character(Animal)),hjust=-0.3,vjust=0.1)+
  theme_classic() +
  xlab("Breed contributes to functional differences") +
  ylab("")
```
But no functional difference between low and high in term of beta diverities.

```{r}
plot_grid(beta_neutral_nmds_vsplit_plot, beta_neutral_nmds_time_plot, beta_phyloge_nmds_plot, beta_function_nmds_plot)
```

Using hillred() phylogenetic and functional redundancy estimations can be carried out relying on the following increasing form of an exponential decay function: y=a*2^(-x/b)+c, where: a is the range of the functional Hill numbers (y value) between the lowest value in the set samples or population and the saturation value b is the units of neutral Hill numbers required to increase half of the range of functional Hill numbers, and the main parameter of interest. c is the value of the saturation point of functional Hill numbers (y value).

As b is dependent of the maximum neutral diversity, the redundancy metric is derived from the division between the estimated parameter b and the maximum neutral diversity value in the sample set.

The function outputs a table with the redundancy value and the estimated parameters, which can be used to plot redundancy charts.

```{r}
high_samp <- rownames(metadata %>%
  filter(Emission == "high" & rownames(metadata) != "D67T5")) 

low_sampl <- rownames(metadata %>%
  filter(Emission == "low" & rownames(metadata) != "D55T5"))

high_samp_count <- count_table_cov[good.mags,high_samp]
low_samp_count <- count_table_cov[good.mags,low_samp]

function_high <- as.data.frame(GIFTs_elements_community) %>% 
  rownames_to_column("Sample") %>% 
  filter(Sample %in% colnames(high_samp_count)) %>% 
  column_to_rownames("Sample")

dist_high <- function_high %>%
  traits2dist(., method = "gower")

#Create the uniqueness matrix
HIGH <- uniqueness(t(function_high), dis = dist_high)

R_HIGH <- HIGH %>% mutate(redundancy = 1-Ui)


function_low <- as.data.frame(GIFTs_elements_community) %>% 
  rownames_to_column("Sample") %>% 
  filter(Sample %in% colnames(low_samp_count)) %>% 
  column_to_rownames("Sample")

dist_low <- function_low %>%
  traits2dist(., method = "gower")

LOW <- funrar::uniqueness(t(function_low), dis = dist_low)

#Redundancy 
R_LOW <- LOW %>% mutate(redundancy = 1-Ui)


F_r <- rbind(R_HIGH, R_LOW) %>% left_join(metadata %>% rownames_to_column("Sample_id"), by = join_by(species == Sample_id)) %>% 
  mutate(Emission = factor(Emission)) %>% 
  left_join(neutral, by = join_by(species == Sample))

#Declare relationship function
relationship_function <- function(x, a, b, c) {return(-a*2^(-x/b)+c)}

ggplot(F_r, aes(x=neutral, y=redundancy, colour=Emission, label = Time)) +
  geom_point(size = 3) +
  #geom_text(hjust=0.6, vjust=1.3, size = 3) +
  geom_smooth(method = "nls", formula = y ~ relationship_function(x, a, b, c), method.args = list(start = redundancy[,-1]), se = FALSE)+
  geom_text_repel(data = F_r, max.overlaps = 10, force = 2, show.legend = F)+
  scale_colour_manual(values=c('#FFD524', '#5c7f9d')) +
  theme(panel.background = element_blank(),
        legend.text = element_text(size=10),
        legend.title = element_text(face='bold',
                                    size=10.5),
        legend.background = element_blank(),
        axis.line = element_line(),
        legend.key = element_blank(),
        legend.position = 'right') +
  xlab('Neutral alpha diversity') +
  ylab('Functional redundancy') +
  guides(col=guide_legend('Emission'))

```



