---
title: "HMSC"
author: "Wanxin_Lai"
date: "2024-05-12"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup}
load("./01_Filter_count.RData")
load("./03_da.RData")
load("./14September2024.RData")
```

```{r , message = FALSE, warning=FALSE}
library(tidyverse)
library(scales)
library(phytools)
library(ggnewscale)
library(DT)
library(cowplot)
library(ggpubr)
library(Hmsc)
library(vioplot)
library(colorspace)
library(ggtree)
library(ggtreeExtra)
library(ape)
library(reshape)
library(distillR)
library(hilldiv2)
```


```{r random effect}
# Random effects data (study design)
StudyDesign <- metadata %>%
  as.data.frame() %>% 
  rownames_to_column("Sample") %>% 
  select(Sample, Animal) %>%
  filter(! Sample %in% c("D55T5", "D67T5")) %>% 
  mutate(Animal = as.factor(Animal)) %>% 
  distinct() %>%
  column_to_rownames("Sample")

# Genome phylogeny
# create ultrametric tree
PData <- read.tree("./mashtree.newick") %>%
  keep.tip(rownames(count_table_cov)) %>% 
  force.ultrametric(.,method="extend")

# Genome count table (quantitative community data)
YData <- count_table_cov %>%
  select(-c("D55T5", "D67T5")) %>% 
  subset(., select = -c("D55T5", "D67T5")) %>% 
  .[PData$tip.label,] %>% 
  mutate(across(where(is.numeric), ~ . +1 )) %>% #add +1 pseudocount to remove zeros
  mutate(across(where(is.numeric), ~ . / (mags_details$Genome_Size / 150) )) %>% #transform to genome counts
  mutate(across(where(is.numeric), ~  log(.) )) %>% #log-transform
  t() # transpose


# Fixed effects data (explanatory variables)
XData <- metadata %>%
  as.data.frame() %>% 
  rownames_to_column("Sample") %>% 
  filter(! Sample %in% c("D55T5", "D67T5")) %>% 
  select(Sample, Time, Breed, Treatment, vsplit_old, Emission) %>%
  mutate(Emission = factor(Emission)) %>%
  mutate(Breed = factor(Breed)) %>%
  mutate(Time = factor(Time)) %>%
  mutate(Treatment = factor(Treatment)) %>% 
  mutate(Vsplit = factor(vsplit_old)) %>% 
  mutate(logseqdepth = count_table_cov %>% #total log-sequencing depth
           select(all_of(row.names(StudyDesign))) %>%
                        colSums() %>%
                        log()
                    ) %>% 
  distinct() %>%
  column_to_rownames("Sample") %>% 
  select(Time, Breed, Treatment, Vsplit, logseqdepth, Emission)

save(YData,XData,StudyDesign,PData,
     file="./hmsc_input_less2.Rdata")
```

### Define formulas of the Hmsc model
# +++ include log sequencing depth
```{r define hmsc model }

# Fixed effects formula
XFormula1 = ~Emission + Breed + logseqdepth + Vsplit + Time

# Study design
rL.animal = HmscRandomLevel(units = levels(StudyDesign$Animal))


#Define models
model1 = Hmsc(Y= YData,
         XData = XData,
         XFormula = XFormula1,
         studyDesign = StudyDesign,
         phyloTree = PData,
         ranLevels = list("Animal"=rL.animal),
         distr = "normal",
         YScale = TRUE)

#Save list of models as an R object.
models = list(model1)#, model2)
if (!dir.exists("modelsminus2")){dir.create("modelsminus2")}
save(models, file = "./modelsminus2/hmsc_models_setup.Rdata")

```

```{r}
if (!dir.exists("modelsminus2/panels")){dir.create("modelsminus2/panels")}

localDir = "./modelsminus2"
ModelDir = file.path(localDir, "models")
PanelDir = file.path(localDir, "panels")
nst = length(thin_list)
```

## Skip this part if you wanna rerun, start from AIC
### Define MCMC
#### Sampling and Data
#### include a threshold at least the mag in minimum 4/5 smaples and relative abundance
```{r declare parameter, eval=FALSE}
models=list(model1)
models=list(model2)
modelnames=c("WithTreatment","NoTreatment")

# How often to sample the MCMC
samples_list = c(50, 150)
# The number of MCMC steps between each recording sample
thin_list = c(1, 5)
# No. of MCMC chain to use
nChains = 4
```

```{r converging, eval=FALSE}
for(Lst in 1:length(samples_list)){
  thin = thin_list[Lst]
  samples = samples_list[Lst]
  for(i in 1:length(models)){
    
    # model fitting
    m = sampleMcmc(models[[i]], 
                   samples = samples,
                   thin=thin,
                   adaptNf=rep(ceiling(0.4*samples*thin),models[[i]]$nr),
                   transient = ceiling(0.5*samples*thin),
                   nChains = nChains,
                   nParallel = nChains)
    
    # save file
    filename=paste("model_",modelnames[i], "_thin_", as.character(thin),
                   "_samples_", as.character(samples),
                   "_chains_",as.character(nChains),".Rdata",sep = "")
    save(m,file=file.path(getwd(),'./modelsminus2',filename))
  }
} 

```

```{r eval = FALSE}

# assess chain convergence

for(i in 1:length(modelnames)){

  ma = NULL
  na = NULL
  ma_gamma = NULL
  na_gamma = NULL
  ma_rho = NULL
  na_rho = NULL
  
  for (Lst in 1:nst) {
    
    thin = thin_list[Lst]
    samples = samples_list[Lst]
    
    filename = paste("model_",modelnames[i], "_thin_", as.character(thin),
                     "_samples_", as.character(samples),
                     "_chains_",as.character(nChains),'.Rdata',sep = "")
    
    #print(filename)
    
    load(file=file.path(localDir,filename))
    
   # print(file)
    
    mpost <- convertToCodaObject(m, spNamesNumbers = c(T,F), covNamesNumbers = c(T,F))
    
    #fixed effects
    psrf.beta <- gelman.diag(mpost$Beta,multivariate=FALSE)$psrf
    
    #traits
    psrf.gamma <- gelman.diag(mpost$Gamma,multivariate=FALSE)$psrf
    
    #phylogeny
    psrf.rho <- gelman.diag(mpost$Rho,multivariate=FALSE)$psrf
    
  
    if(is.null(ma)){
      ma=psrf.beta[,1]
      na = paste0("model_",as.character(i),"_",as.character(thin),",",as.character(samples))
    } else {
      ma = cbind(ma,psrf.beta[,1])
      na = c(na,paste0("model_",as.character(i),"_",as.character(thin),",",as.character(samples)))
    }
    if(is.null(ma_gamma)){
      ma_gamma=psrf.gamma[,1]
      na_gamma = paste0("model_",as.character(i),"_",as.character(thin),",",as.character(samples))
    } else {
      ma_gamma = cbind(ma_gamma,psrf.gamma[,1])
      na_gamma = c(na_gamma,paste0("model_",as.character(i),"_",as.character(thin),",",as.character(samples)))
    }
    if(is.null(ma_gamma)){
      ma_rho=psrf.rho[,1]
      na_rho = paste0("model_",as.character(i),"_",as.character(thin),",",as.character(samples))
    } else {
      ma_rho = cbind(ma_rho,psrf.rho[,1])
      na_rho = c(na_rho,paste0("model_",as.character(i),"_",as.character(thin),",",as.character(samples)))
    }
  }
  panel.name=paste("MCMC_convergence", "_model_",as.character(i),".pdf",sep = "")
  pdf(file=file.path(PanelDir,panel.name))
  #print(file)
  par(mfrow=c(2,1))
  vioplot(ma,col=rainbow_hcl(1),names=na,ylim=c(0,max(ma)),main="psrf(beta)")
  vioplot(ma,col=rainbow_hcl(1),names=na,ylim=c(0.9,1.1),main="psrf(beta)")
  par(mfrow=c(2,1))
  vioplot(ma_gamma,col=rainbow_hcl(1),names=na_gamma,ylim=c(0,max(ma_gamma)),main="psrf(gamma)")
  vioplot(ma_gamma,col=rainbow_hcl(1),names=na_gamma,ylim=c(0.9,1.1),main="psrf(gamma)")
  par(mfrow=c(2,1))
  boxplot(ma_rho,col=rainbow_hcl(1),names=na_rho,ylim=c(0,max(ma_rho)),main="psrf(rho)")
  dev.off()
}
```

To test for MCMC convergence, measuring the scale reduction factor: the beta (response to perturbations), gamma (trait response to perturbations), and rho parameters (phylogenetic signal).


### Model variance
```{r eval=FALSE}
preds <- computePredictedValues(m)

MF = evaluateModelFit(hM=m, predY=preds)
mean(MF$R2, na.rm = T)
```



### cross validation 80:20 AUC
### AIC
```{r eval =FALSE}
modelnames=c("NoTreatment")

####Evaluate model fit

i=1 #Notime
filename = paste("model_",modelnames[i], "_thin_", as.character(thin),
                 "_samples_", as.character(samples),
                 "_chains_",as.character(nChains),".Rdata",sep = "")
print(filename)
load(file=file.path(localDir,filename))

m_NoInt=m
m_NoInt_WAIC=computeWAIC(m_NoInt,byColumn =TRUE)
median(m_NoInt_WAIC,na.rm = TRUE)
predY.MF.NoInt=computePredictedValues(m_NoInt, expected=FALSE)
EpredY.MF.NoInt=apply(predY.MF.NoInt, MARGIN=1:2, mean)

plot(m_NoInt$Y,EpredY.MF.NoInt)
abline(coef = c(0,1),col="red")
MF.NoInt=evaluateModelFit(hM=m_NoInt, predY=predY.MF.NoInt)
mean(MF.NoInt$R2)
```

```{r eval = FALSE}
i=2 
filename = paste("model_",modelnames[i], "_thin_", as.character(thin),
                 "_samples_", as.character(samples),
                 "_chains_",as.character(nChains),".Rdata",sep = "")
load(file=file.path(localDir,filename))
print(filename)
m_NoInt=m
m_NoInt_WAIC=computeWAIC(m_NoInt,byColumn =TRUE)
median(m_NoInt_WAIC,na.rm = TRUE)
predY.MF.NoInt=computePredictedValues(m_NoInt, expected=FALSE)
EpredY.MF.NoInt=apply(predY.MF.NoInt, MARGIN=1:2, mean)

plot(m_NoInt$Y,EpredY.MF.NoInt)
abline(coef = c(0,1),col="red")
MF.NoInt=evaluateModelFit(hM=m_NoInt, predY=predY.MF.NoInt)
mean(MF.NoInt$R2)
```


A support threshold of ‘0.05’ to assign significance, meaning that parameters with posterior overlaps of <10% are considered significantly different.

```{r}

# Select desired support threshold
support=0.9
negsupport=1-support

# Basal tree
postestimates_tree <- tree %>%
        keep.tip(., tip=m$spNames)

# Posterior estimate table
post_beta <- getPostEstimate(hM=m, parName="Beta")$support %>%
    as.data.frame() %>%
    mutate(variable=m$covNames) %>%
    pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
    mutate(genome=factor(genome, levels=rev(postestimates_tree$tip.label))) %>%
    #mutate(value = case_when(
    #      value >= support ~ "Positive",
    #      value <= negsupport ~ "Negative",
    #      TRUE ~ "Neutral")) %>%
    mutate(value=factor(value, levels=c("Positive","Neutral","Negative"))) %>%
    pivot_wider(names_from = variable, values_from = value) %>%
    ##rename(intercept=2) %>%
    ##select(genome,dominance,groupvariable) %>%
    column_to_rownames(var="genome")

```


```{r}
# Basal ggtree
postestimates_tree <- postestimates_tree %>%
        force.ultrametric(.,method="extend") %>%
        ggtree(., size = 0.3) #+ layout_dendrogram()

phylum_colour <- data.frame(Phylum = (unique(mags_phylum_genomes_sorted$Phylum)), color = (hue_pal()(12)))

Phylum_colour <- setNames(phylum_colour$color, phylum_colour$Phylum)

heatmap <- mags_phylum_genomes_sorted %>% arrange(match(Genome, postestimates_tree$tip.label)) %>%
  select(Genome,Phylum) %>%
    mutate(Phylum = factor(Phylum, levels = unique(Phylum))) %>%
    column_to_rownames(var = "Genome")

postestimates_tree <- gheatmap(postestimates_tree, heatmap, offset=0, width=0.1, colnames=FALSE) +
      labs(fill="Phylum") 

```


```{r}
#Reset fill scale to use a different colour profile in the heatmap
postestimates_tree <- postestimates_tree + new_scale_fill()

# Add posterior significant heatmap
postestimates_tree <- gheatmap(postestimates_tree, post_beta, offset=0.1, width=0.3, colnames=TRUE, colnames_position="top",colnames_angle=90, colnames_offset_y=0.8, hjust=0, font.size = 1.5) +
        scale_fill_manual(values=c("#be3e2b","#f4f4f4","#b2b530"))+
        labs(fill="Trend")+ theme(legend.position = "none")


#Reset fill scale to use a different colour profile in the heatmap
postestimates_tree <- postestimates_tree + new_scale_fill()

```

```{r}
name_list <- tibble(full_name = colnames(GIFTs_elements_names)) %>%
  separate(full_name, into = c("Code", "Elements"), sep = " ", extra = "merge") %>% 
  mutate(Full_name = colnames(GIFTs_elements_names)) %>% 
  select(!Elements) 

# Create a named vector for renaming
function_table_name <- GIFTs_elements_filtered %>% 
  rename_with(~deframe(name_list)[.x], .cols = name_list$Code) %>% 
  select(any_of(name_list$Full_name))

function_table <- function_table_name %>% 
  rownames_to_column(var = "Genome") %>%
arrange(match(Genome, postestimates_tree$tip.label)) %>% 
  arrange(match(Genome, tree$tip.label)) %>% 
  column_to_rownames("Genome")
```


```{r fig.height=11}

# Select desired support threshold
support=0.9
negsupport=1-support

postestimate_test_emi <- getPostEstimate(hM=m, parName="Beta")$support %>%
      as.data.frame() %>%
      mutate(variable=m$covNames) %>%
      pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
      mutate(trend = case_when(
            value >= support ~ "Positive",
            value <= negsupport ~ "Negative",
            TRUE ~ "Neutral")) %>%
      filter(variable==c("Emissionlow"), trend!="Neutral") %>%
      select(genome,trend) %>%
      left_join(function_table %>% rownames_to_column(var="genome"), by=join_by(genome==genome)) %>%
      pivot_longer(-c(genome,trend), names_to = "trait", values_to = "value") %>%
      group_by(trait) %>%
      summarise(p_value = wilcox.test(value ~ trend)$p.value) %>%
      mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
      filter(p_adjust < 0.05)


postestimate_test_breed <- getPostEstimate(hM=m, parName="Beta")$support %>%
      as.data.frame() %>%
      mutate(variable=m$covNames) %>%
      pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
      mutate(trend = case_when(
            value >= support ~ "Positive",
            value <= negsupport ~ "Negative",
            TRUE ~ "Neutral")) %>%
      filter(variable==c("BreedLuing"), trend!="Neutral") %>%
      select(genome,trend) %>%
      left_join(function_table %>% rownames_to_column(var="genome"), by=join_by(genome==genome)) %>%
      pivot_longer(-c(genome,trend), names_to = "trait", values_to = "value") %>%
      group_by(trait) %>%
      summarise(p_value = wilcox.test(value ~ trend)$p.value) %>%
      mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
      filter(p_adjust < 0.05)


# extra = "merge" argument ensures that any additional spaces in the names are included in the last_name column.
name_list <- tibble(full_name = colnames(GIFTs_elements_names)) %>%
  separate(full_name, into = c("Code", "Elements"), sep = " ", extra = "merge") %>% 
  mutate(Full_name = colnames(GIFTs_elements_names)) %>% 
  select(!Elements) %>% 
  left_join(postestimate_test_emi, by = join_by(Full_name == trait)) %>% 
  mutate(new_Full_name = ifelse(!is.na(p_adjust),paste(Full_name,"*", sep = " "), Full_name))# %>% 


name_list_selected <- name_list %>% select(Code, new_Full_name)
                                           
# Convert the selected name_list to a named vector using deframe
name_vector <- deframe(name_list_selected)


function_table_name <- GIFTs_elements_filtered %>%
  rename_with(~ name_vector[.x], .cols = names(name_vector)) %>%
  select(any_of(name_list_selected$new_Full_name))


function_table_postestimate <- function_table_name %>% 
  rownames_to_column(var = "Genome") %>%
arrange(match(Genome, postestimates_tree$tip.label)) %>% 
  arrange(match(Genome, tree$tip.label)) %>% 
  column_to_rownames("Genome")
 
hclust_func_community <- hclust(dist(t(function_table_postestimate)), method = "ward.D2")

```

```{r}
postestimates_tree_plot <- gheatmap(postestimates_tree, function_table_postestimate[,hclust_func_community$order], offset=0.42, width = 4, colnames=TRUE, colnames_angle = 90, hjust = 0, colnames_position = "top", font.size = 1.5) +
            vexpand(.1) +
            coord_cartesian(clip = "off") +
            scale_fill_gradient(low = "darkorange", high = rev(c("#4C00FF","#00E5FF","yellow")),na.value="white") +
            labs(fill="Function fullness")

postestimates_tree_plot

noleg <- postestimates_tree_plot + theme(legend.position = "none")

noleg 
```

```{r}
breakList = seq(0, 1, length.out = 101)
#colors <- colorRampPalette(c("orange","yellow","#00E5FF","#4C00FF"))(100)
annot.col <- data.frame(row.names = count_table$Genome, Phylum = as.factor(count_table$Phylum))

phylum_colour <- data.frame(Phylum = (unique(count_table$Phylum)), color = (hue_pal()(12)))
Phylum_colour <- setNames(phylum_colour$color, phylum_colour$Phylum)

pheatmap(t(function_table_postestimate),
         cluster_rows = T,
         cluster_cols = T,
         #color = colors,
         cutree_rows = 5,
         cutree_cols = 2,
         fontsize_row= 4.6,
         annotation_col = annot.col,
         show_colnames = F,
         clustering_method="ward.D2",
         annotation_colors = list(Phylum = Phylum_colour)
         #annotation_legend = 0
         )
```

### MAGs positively associated with low methane emission

```{r}
low.emi.MAG <- getPostEstimate(hM=m, parName="Beta")$support %>%
    as.data.frame() %>%
    mutate(variable=m$covNames) %>%
    pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
    mutate(trend = case_when(
          value >= support ~ "Positive",
          value <= negsupport ~ "Negative",
          TRUE ~ "Neutral")) %>%
    filter(variable=="Emissionlow", trend=="Positive") %>%
    arrange(-value) %>%
    left_join(mags_details,by=join_by(genome==Genome)) 
write.csv(low.emi.MAG, file = "./modelsminus2/panels/low_emi_MAGs.csv")
datatable(low.emi.MAG)
```

### MAGs negatively associated with low methane emission

```{r}
high.emi.MAG <- getPostEstimate(hM=m, parName="Beta")$support %>%
    as.data.frame() %>%
    mutate(variable=m$covNames) %>%
    pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
    mutate(trend = case_when(
          value >= support ~ "Positive",
          value <= negsupport ~ "Negative",
          TRUE ~ "Neutral")) %>%
    filter(variable=="Emissionlow", trend=="Negative") %>%
    #mutate(value = 1- value) %>% 
    arrange(-value) %>%
    left_join(mags_details,by=join_by(genome==Genome))
high.emi.MAG
write.csv(high.emi.MAG, file = "~/Small_Cow/MAG/hmsc_split_seaweed/modelsminus2/panels/high_emi_MAGs.csv")
datatable(high.emi.MAG)

#save(low.emi.MAG,high.emi.MAG, file = "~/Small_Cow/MAG/hmsc_split_seaweed/panels/emi_lowhigh_MAGs_hmsc.RData")
```


### Estimate - support plot (Emission)  
```{r}
estimate <- getPostEstimate(hM=m, parName="Beta")$mean %>%
    as.data.frame() %>%
    mutate(variable=m$covNames) %>%
    filter(variable=="Emissionlow") %>%
    pivot_longer(!variable, names_to = "genome", values_to = "mean") %>%
    select(genome,mean)

support <- getPostEstimate(hM=m, parName="Beta")$support %>%
    as.data.frame() %>%
    mutate(variable=m$covNames) %>%
    filter(variable=="Emissionlow") %>%
    pivot_longer(!variable, names_to = "genome", values_to = "support") %>%
    select(genome,support)

inner_join(estimate,support,by=join_by(genome==genome)) %>%
    mutate(significance=ifelse(support >= 0.9 | support <= 0.1,1,0)) %>%
    mutate(support=ifelse(mean<0,1-support,support)) %>%
    left_join(mags_phylum_genomes_sorted, by = join_by(genome == Genome)) %>%
    mutate(phylum = ifelse(support > 0.9, Phylum, NA)) %>%
    ggplot(aes(x=mean,y=support, color = phylum))+
      geom_point(alpha=0.7, shape=16, size=3)+
      scale_color_manual(values = Phylum_colour) +
      geom_vline(xintercept = 0) +
      xlim(c(-0.4,0.4)) +
      labs(x = "Beta regression coefficient", y = "Posterior probability") +
      theme_minimal()+
       theme(legend.position = "right")
```

### Dispersion plot for Emission
```{r}
gradient = c("Emissionlow", "Emissionhigh") #vector value ,can specific by timepoint

gradientlength = length(gradient)

#Treatment-specific gradient predictions
pred <- constructGradient(m, 
                      focalVariable = "Emission", #when other variables are not involved
                      ngrid=gradientlength) %>%
             predict(m, Gradient = ., expected = TRUE)


#Treatment-specific gradient predictions
predY <- pred %>%
        do.call(rbind,.) %>%
        as.data.frame() %>%
        mutate(Emission=rep(gradient,600)) %>%
        pivot_longer(-c(Emission), names_to = "genome", values_to = "value") %>% 
  left_join(mags_phylum_genomes_sorted, by = join_by(genome == Genome)) 

predY.emi <- predY

# Select desired support threshold
support=0.9
negsupport=1-support



test.emi <- getPostEstimate(hM=m, parName="Beta")$support %>%
    as.data.frame() %>%
    mutate(variable=m$covNames) %>%
    pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
    mutate(trend = case_when(
          value >= support ~ "Positive",
          value <= negsupport ~ "Negative",
          TRUE ~ "Neutral")) %>%
    filter(variable=="Emissionlow") %>%
    select(genome,trend,value)

colnames(predY)[colnames(predY) == "value"] <- "predicted_value"
colnames(test.emi)[colnames(test.emi) == "value"] <- "stat_value"

test.emi <- test.emi %>% 
    left_join(predY, by=join_by(genome==genome)) %>%
    #filter(trend != "Neutral") %>%
    #filter(genome %in% predY_asc)  %>% #only display genomes with contrasting dynamics across treatments
    group_by(genome, trend, Emission)

test.emi <- test.emi %>% 
    summarize(predicted_value = mean(predicted_value, na.rm = TRUE), stat_value = mean(stat_value, na.rm = TRUE)) %>%
    left_join(mags_phylum_genomes_sorted, by=join_by(genome == Genome)) 

test.emi <- test.emi[order(-Emission, genome)] 

test.emi %>%
    ggplot(aes(x=Emission, y=predicted_value, group=genome, color=Phylum, linetype=trend)) + 
        geom_line() +
        scale_linetype_manual(values=c("solid","dashed","solid")) +
        #scale_color_manual(values=phylum_colors) +
        facet_grid(fct_rev(trend) ~ Phylum) +
        labs(y="Genome abundance (log)",x="Emission") +
        theme(legend.position = "none") +
        theme_minimal() +
         theme(legend.position = "none",
               strip.text.x = element_text(size = 7), 
               axis.text.x = element_text(angle = 45, hjust = 0.8, size = 7),
               axis.line.x = element_line(size = 0.5, linetype = "solid", colour = "black"),
)
```

```{r}


# Refernece tree (for sorting genomes)
genome_tree_subset <- tree %>%
        keep.tip(., tip=m$spNames) 

predY.emi %>%
      pivot_wider(names_from = Emission, values_from = value) %>%
      unnest(c(`Emissionlow`, `Emissionhigh`)) %>%
      mutate(diff_low.high = `Emissionlow` - `Emissionhigh`) %>%
      select(genome, diff_low.high) %>%
      left_join(mags_phylum_genomes_sorted, by=join_by(genome == Genome)) %>%
      mutate(genome= factor(genome, levels=genome_tree_subset$tip.label)) %>%
      ggplot(., aes(y=genome, x=diff_low.high, fill=Phylum, color=Phylum)) +
          #scale_color_manual(values=colors_alphabetic)+
          geom_vline(xintercept = 0)+
          #scale_fill_manual(values=paste0(colors_alphabetic,"80"))+
          geom_boxplot(outlier.shape = NA) +
          theme_classic() +
          theme(axis.text.y = element_blank())
```




```{r}
#Compute the residual correlation matrix
OmegaCor <- computeAssociations(m)

# Refernece tree (for sorting genomes)
genome_tree_subset <- tree %>%
        keep.tip(., tip=m$spNames) 


#Co-occurrence matrix at the animal level
supportLevel = 0.95
toPlot = ((OmegaCor[[1]]$support>supportLevel)
          + (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean

phylum_colour <- data.frame(Phylum = (unique(count_table$Phylum)), color = (hue_pal()(12)))
Phylum_colour <- setNames(phylum_colour$color, phylum_colour$Phylum)
annot.row <- data.frame(row.names = count_table$Genome, Phylum = as.factor(count_table$Phylum))

pheatmap(mat = toPlot, 
        treeheight_row = 0, 
         treeheight_col = 0,
         show_rownames = F, 
         show_colnames = F,
         annotation_col = annot.row,
         annotation_colors = list(Phylum = Phylum_colour),
         annotation_legend = 0)

```


```{r eval = FALSE, include = FALSE}

# Get upper triangle of the correlation matrix
  get_upper_tri <- function(cormat){
    cormat[lower.tri(cormat)]<- NA
    return(cormat)
  }
  
toPlot_upper <- get_upper_tri(toPlot)

melted_eSim_upper <- melt(toPlot_upper, na.rm = TRUE) %>%
  arrange(value)

fac_levels <- melted_eSim_upper$X1 %>% unique()

data_used <- melted_eSim_upper %>% 
  mutate(genome1 = factor(melted_eSim_upper$X1, levels = fac_levels)) %>% 
  mutate(genome2 = factor(melted_eSim_upper$X2, levels = rev(fac_levels)))

p <- data_used %>% ggplot(aes(x = genome2, y = genome1, fill = value)) 
p <- p + geom_tile() 
p <- p + scale_fill_gradient2(low = "#be3e2b", mid = "#f4f4f4", high = "#b2b530")
p <- p + theme(text=element_text(size=3, family="serif"))
p <- p + theme_void()


p

vtree <- genome_tree_subset %>%
  force.ultrametric(.,method="extend") %>%
  ggtree(.)

genome_tree_subset$tip.label


matrix2 <- toPlot %>%
  as.data.frame() %>% 
  rownames_to_column(var="genome1") %>%
  pivot_longer(!genome1, names_to = "genome2", values_to = "cor") %>%
  mutate(genome1= factor(genome1, levels=genome_tree_subset$tip.label)) %>%
  mutate(genome2= factor(genome2, levels=genome_tree_subset$tip.label)) %>% 
  ggplot(aes(x = genome1, y = genome2, fill = cor)) +
            geom_tile() + 
            scale_fill_gradient2(low = "#be3e2b",
                       mid = "#f4f4f4",
                       high = "#b2b530")+
            theme_void()

vtree <- genome_tree_subset %>%
  force.ultrametric(.,method="extend") %>%
  ggtree(.)

corr_phylum_tree <- gheatmap(vtree, heatmap, offset=0, width=0.05, colnames=FALSE) +
      labs(fill="Phylum") +
   theme(legend.position = "none") 
  
```

```{r fig.width= 10, fig.height=10, eval = FALSE, include = FALSE}
#create composite figure
leg1 <- get_legend(gheatmap(vtree, heatmap, offset=0, width=0.05, colnames=FALSE, legend_title = "Phylum") + labs(fill="Phylum") +
   theme(legend.position = "right"))

top_row <- plot_grid(corr_phylum_tree, matrix, align = "h", ncol = 2, rel_heights = c(1, 0.25), rel_widths = c(0.4, 1))

plot_grid(top_row, leg1, ncol = 2, rel_widths = c(1,0.2))
```
 
```{r}
# Compute variance partitioning
varpart=computeVariancePartitioning(m)

datatable(varpart$vals %>%
   as.data.frame() %>%
   rownames_to_column(var="variable") %>%
   pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
   mutate(variable=factor(variable)) %>%
   group_by(variable) %>%
   summarise(mean=mean(value)*100,sd=sd(value)*100))


```

```{r}
# Basal tree
varpart_tree <- tree %>%
        keep.tip(., tip=m$spNames) 

#Varpart table
varpart_table <- varpart$vals %>%
   as.data.frame() %>%
   rownames_to_column(var="variable") %>%
   pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
   mutate(variable=factor(variable)) %>%
   mutate(genome=factor(genome, levels=rev(varpart_tree$tip.label)))

# Basal ggtree
varpart_tree <- varpart_tree %>%
        force.ultrametric(.,method="extend") %>%
        ggtree(., size = 0.3)
```

```{r}
# Add phylum colors next to the tree tips
heatmap <- mags_phylum_genomes_sorted %>% arrange(match(Genome, varpart_tree$tip.label)) %>%
  select(Genome,Phylum) %>%
    mutate(Phylum = factor(Phylum, levels = unique(Phylum))) %>%
    column_to_rownames(var = "Genome")

varpart_tree <- gheatmap(varpart_tree, heatmap, offset=0, width=0.1, colnames=FALSE) +
      labs(fill="Phylum")

#Reset fill scale to use a different colour profile in the heatmap
varpart_tree <- varpart_tree + new_scale_fill()

# Add variance stacked barplot
vertical_tree <-  varpart_tree +
        #scale_fill_manual(values=c("#34738f","#cccccc","#ed8a45","#b2b530","#be3e2b","#f6de6c","#83bb90"))+
        scale_fill_manual(values=c("#34738f","#cccccc","#ed8a45","#b2b530","#be3e2b","#83bb90","#f6de6c"))+ #"#122f3d"
        geom_fruit(
             data=varpart_table,
             geom=geom_bar,
             mapping = aes(x=value, y=genome, fill=variable, group=variable),
             pwidth = 2,
             offset = 0.17,
             width= 1,
             orientation="y",
             stat="identity")+
      labs(fill="Variable")

vertical_tree

```

## Community functional prediction

### Plot traits linked to trends (metabolites vs "Emission")
#### Testing functional trait differences
Using Wilcoxon test with Benjamini-Hochberg procedure for FDR, identify functional traits that differ between genomes that are either positively or negatively association with emission.
```{r}
# Select desired support threshold
support=0.9
negsupport=1-support

postestimate_test_emi <- getPostEstimate(hM=m, parName="Beta")$support %>%
      as.data.frame() %>%
      mutate(variable=m$covNames) %>%
      pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
      mutate(trend = case_when(
            value >= support ~ "Positive",
            value <= negsupport ~ "Negative",
            TRUE ~ "Neutral")) %>%
      filter(variable==c("Emissionlow"), trend!="Neutral") %>%
      select(genome,trend) %>%
      left_join(function_table %>% rownames_to_column(var="genome"), by=join_by(genome==genome)) %>%
      pivot_longer(-c(genome,trend), names_to = "trait", values_to = "value") %>%
      group_by(trait) %>%
      summarise(p_value = wilcox.test(value ~ trend)$p.value) %>%
      mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
      filter(p_adjust < 0.05)

datatable(postestimate_test_emi)
write.csv(postestimate_test_emi, file = "~/Small_Cow/MAG/hmsc_split_seaweed/modelsminus2/panels/Fucntionality_differences_MAG.csv")

```





### Plot traits linked to trends (metabolites vs "vsplit")
```{r}
# Select desired support threshold
support=0.9
negsupport=1-support

postestimate_test_vsplit <- getPostEstimate(hM=m, parName="Beta")$support %>%
      as.data.frame() %>%
      mutate(variable=m$covNames) %>%
      pivot_longer(!variable, names_to = "genome", values_to = "value") %>%
      mutate(trend = case_when(
            value >= support ~ "Positive",
            value <= negsupport ~ "Negative",
            TRUE ~ "Neutral")) %>%
      filter(variable=="Vsplitgroup2", trend!="Neutral") %>%
      select(genome,trend) %>%
      left_join(function_table %>% rownames_to_column(var="genome"), by=join_by(genome==genome)) %>%
      pivot_longer(-c(genome,trend), names_to = "trait", values_to = "value") %>%
      group_by(trait) %>%
      summarise(p_value = wilcox.test(value ~ trend)$p.value) %>%
      mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
      filter(p_adjust < 0.05)

datatable(postestimate_test_vsplit)


```


### Element level
```{r}
predY <- pred %>%
        do.call(rbind,.) %>%
        as.data.frame() %>%
        mutate(Emission=rep(gradient,600)) %>%
        pivot_longer(-c(Emission), names_to = "genome", values_to = "value") 

#Aggregate bundle-level GIFTs into the compound level
GIFTs_elements <- to.elements(genome_gifts, gift_cm)

# GIFT at element level(Aggregate bundle-level GIFTs into the compound level)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% rownames(count_table_cov_size %>% 
  select(-c("D55T5", "D67T5"))), ] %>% 
  as.data.frame() %>%  
  select_if(~ sum(.) != 0) #remove column with only 0s

# microbiome-wide GIFTs can also be computed by relying on the slopes of iterations.
#600 iters and 149 elements
community_elements <- predY %>%
  group_by(Emission, genome) %>%
  mutate(row_id = row_number()) %>%
  pivot_wider(names_from = genome, values_from = value) %>%
  ungroup() %>%
  group_split(row_id) %>%
  as.list() %>%
  lapply(., FUN = function(x){x %>%
    select(-row_id) %>%
    column_to_rownames(var = "Emission") %>%
    as.data.frame() %>%
    exp() %>%
    t() %>%
      tss() %>% 
    to.community(GIFTs_elements_filtered,.,gift_cm) %>% 
      as.data.frame() %>%
    rownames_to_column(var="Emission")
   })
```

```{r}
element_pred_em <- map_dfr(community_elements, function(df) {
  EL <- df %>% filter(Emission == "Emissionlow") %>% select(-Emission)
  EH <- df %>% filter(Emission == "Emissionhigh") %>% select(-Emission)
  EH - EL
}) %>%
  mutate(iteration=c(1:600)) %>% 
  pivot_longer(!iteration,names_to="trait",values_to="value") %>%
      group_by(trait) %>%
      summarise(mean=mean(value),
        p1 = quantile(value, probs = 0.1),
        p9 = quantile(value, probs = 0.9),
        positive_support = sum(value > 0)/600,
        negative_support = sum(value < 0)/600) %>%
      arrange(-positive_support)

# Positively associated
element_pred_em %>%
  filter(mean >0) %>%
  arrange(-positive_support) %>%
  filter(positive_support>=0.9) 

# Negatively associated
element_pred_em %>%
  filter(mean <0) %>%
  arrange(-negative_support) %>%
  filter(negative_support>=0.9) 
```

```{r}
positive <- element_pred_em %>%
  filter(mean >0) %>%
  arrange(mean) %>%
  filter(positive_support>=0.9) %>%
  left_join(name_list, by = join_by(trait == Code)) %>% 
  dplyr::select(-negative_support) %>%
  dplyr::rename(support=positive_support)

negative <- element_pred_em %>%
  filter(mean <0) %>%
  arrange(mean) %>%
  filter(negative_support>=0.9) %>%
  left_join(name_list, by = join_by(trait == Code)) %>% 
  dplyr::select(-positive_support) %>%
  dplyr::rename(support=negative_support)

```


```{r}
calculate_slope <- function(x) {
  y <- unlist(x)
  
  x_seq <- seq_along(y)
  
  # Fit a linear model
  lm_fit <- lm(y ~ x_seq)
  
  slope <- coef(lm_fit)[2]
  
  return(slope)
}
```

```{r}
map_result <- map_dfc(community_elements, function(mat) {
      mat %>%
        column_to_rownames(var = "Emission") %>%
        t() %>%
        as.data.frame() %>%
        rowwise() %>%
        mutate(slope = calculate_slope(c_across(everything()))) %>%
        select(slope) })

map_result_cols_named <- map_result %>%
      t() %>%
      as.data.frame() %>%
      set_names(colnames(community_elements[[1]])[-1])


element_predictions <- map_result_cols_named %>%
      rownames_to_column(var="iteration") %>%
      pivot_longer(!iteration, names_to="trait",values_to="value") %>%
      group_by(trait) %>%
       summarise(mean=mean(value),#trimming the top 10% extreme values
                median=median(value),
        p1 = quantile(value, probs = 0.1),
        p9 = quantile(value, probs = 0.9),
        positive_support = sum(value > 0)/600,
        negative_support = sum(value < 0)/600) %>%
      arrange(-positive_support)

# Positively associated
datatable(element_predictions %>%
            left_join(name_list, by = join_by(trait == Code)) %>% 
  filter(mean >0) %>%
  arrange(-positive_support))
```

```{r}
# Negatively associated
datatable(element_predictions %>%
            left_join(name_list, by = join_by(trait == Code)) %>% 
  filter(mean <0) %>%
  arrange(-negative_support) )
```

```{r}
positive <- element_predictions %>%
  left_join(name_list, by = join_by(trait == Code)) %>% 
  filter(mean >0) %>%
  arrange(-positive_support) %>%
  select(-negative_support) %>%
  dplyr::rename(support=positive_support) %>%
  filter(p_value < 0.05 & support >= 0.7 & p1 >= -0.01)# %>%
  #dplyr::slice(1:30)
  
negative <- element_predictions %>%
  left_join(name_list, by = join_by(trait == Code)) %>% 
 filter(mean <0) %>%
  arrange(-negative_support) %>%
  select(-positive_support) %>%
  dplyr::rename(support=negative_support) %>%
  mutate(support=-support) %>%
  arrange(support) %>%
  filter(p_value < 0.05 & support <= -0.7, p9 <= 0.01) %>% 
  arrange(-mean)

# Define custom transformation
even_spacing_trans <- trans_new(
  name = "even_spacing",
  transform = function(x) {
    breaks <- custom_breaks
    scaled_x <- approx(breaks, seq_along(breaks), xout = x)$y - 1
    return(scaled_x / (length(breaks) - 1))
  },
  inverse = function(x) {
    breaks <- custom_breaks
    scaled_x <- approx(seq_along(breaks), breaks, xout = x * (length(breaks) - 1) + 1)$y
    return(scaled_x)
  },
  domain = range(custom_breaks)
)

padj.element <- postestimate_test[postestimate_test$p_value < 0.05,]

test.elements <- bind_rows(positive,negative) %>%
 # merge(., name_list, by.x="trait", by.y= "Code", all.x = T ) %>%
  left_join(gift_cm %>% select(Code_element, Function) %>% unique(), by = join_by(trait == Code_element)) %>% 
  left_join(postestimate_test %>% select(trait, p_adjust), by = join_by(Full_name == trait))

test.elements$new_Full_name <- gsub("\\*", "", test.elements$new_Full_name)
positive$new_Full_name <- gsub("\\*", "", positive$new_Full_name)
negative$new_Full_name <- gsub("\\*", "", negative$new_Full_name)



test.elements %>% 
  filter(!str_detect(Function, "Cellular structure|Appendages|Spore")) %>% 
  mutate(trait=factor(new_Full_name,levels=c(positive$new_Full_name, negative$new_Full_name))) %>%
  ggplot(aes(x=mean, y=fct_rev(trait), xmin=p1, xmax=p9, color=Function)) +
      geom_point() +
      geom_errorbar() +
      scale_color_manual(values = c(rainbow(14))) +
      theme_minimal() +
  theme(axis.text.y = element_text(size = 12), legend.text = element_text(size = 11.5)) +
  geom_point() +
      labs(x="Regression coefficient",y="Functional trait")
```
